### 2024-03-27 sssomt.tex
Initial commit.

\documentclass{ceurart}
\sloppy
\usepackage{listings}
\lstset{breaklines=true}

\def\species#1{\textit{#1}}
\def\term#1{`#1'}
\def\property#1{\textit{#1}}
\def\curie#1{\textsf{#1}}
\def\function#1{\textit{#1}}

\begin{document}
\copyrightyear{2024}
\copyrightclause{Copyright for this paper by its authors.
  Use permitted under Creative Commons License Attribution 4.0
  International (CC BY 4.0).}

\conference{Proceedings of the International Conference on Biomedical
Ontologies 2024, July 17--19, 2024, Enschede, The Netherlands}

\title{(Re-)bridging the anatomy ontologies with SSSOM}

\author[1]{Damien Goutte-Gattat}[%
orcid=0000-0002-6095-8718,
email=dpg44@cam.ac.uk,
]
\cormark[1]
%\fnmark[1]
\address[1]{University of Cambridge, Downing Street, Cambridge, CB2 3DY, United Kingdom}

\author[2]{Nicolas Matentzoglu}[
orcid=0000-0002-7356-1779,
email=nicolas.matentzoglu@gmail.com,
]
\address[2]{Semanticly, Athens 10563, Greece}

\cortext[1]{Corresponding author.}
%\fntext[1]{These authors contributed equally.}

\begin{abstract}
Ontologies that describe the anatomical structures and cell types of model
organisms are critically required for the annotation and successful
exploitation of high-throughput datasets. The Uberon ontology (for anatomical
structures) and the Cell Ontology (CL, for cell types) jointly aim to provide a
consistent ontology that can be used across a wide range of species, by
leveraging existing species-centric anatomy ontologies to construct a single
integrated multi-species ontology. In this paper, we describe how we overhauled
the integration mechanism between Uberon, CL, and the species-centric
ontologies by using the newly devised SSSOM (``Simple Standard for Sharing
Ontological Mappings'') standard to manage the mappings between all concerned
ontologies.
\end{abstract}

\begin{keywords}
  Anatomy ontologies \sep
  mappings \sep
  cross-species studies
\end{keywords}

\maketitle

\section{Introduction}

One of the pre-requisites for the exploitation of datasets from
high-throughput experimental methods, such as single-cell RNA sequencing
datasets, beyond their lab of origin is the ability to annotate samples
and results with consistent terms. That is, we need controlled
vocabularies to properly record information such as the exact
techniques used, the organs or tissues from which the experimental
samples were taken, or the cell types that were identified. Several
ontologies have been developed for this kind of purpose, such as the
Experimental Factor Ontology (EFO) to annotate experimental
methods~\cite{malone2010}, the Drosophila Anatomy Ontology (FBbt) to
annotate anatomical structures and cell types in
\species{D.~melanogaster}~\cite{costa2013}, the Worm Anatomy Ontology
(WBbt) to do the same in \species{C.~elegans}~\cite{lee2003}, etc. Most
of those ontologies are grouped under the umbrella of the OBO
Foundry~\cite{jackson2021}.

However, comparing datasets across species additionally requires that
the ontology terms used to annotate the datasets are themselves
comparable. For example, comparing a dataset of mouse blood cells with a
dataset of fly hemocytes requires that some link (relationship) exist
between the term representing the concept of blood in mice and the term
representing the concept of hemolymph in flies. Ideally, this would in
turn require that a single cross-species ontology be used for all
datasets, containing all the terms needed to represent all anatomical
structures and cell types with their species-specific variations,
organised in a consistent hierarchy.

\section{The Uberon strategy for a multi-species ontology}

Building a single cross-species anatomy ontology from the ground up, with terms
suitable for all the model organisms, would be a gigantic task, requiring lots
of species-specific expertise that would be hard to gather in a single project.
Instead, a more realistic approach was adopted by the Uberon
project~\cite{mungall2012a} which, rather than attempting to describe the
anatomy of every model organism, leverages the existing species-specific
anatomy ontologies. Uberon aims to provide a ``core'' anatomy ontology that
describes anatomical structures in a species-neutral way, along with
``bridges'' that allow to integrate the more precise terms from the
species-specific ontologies (Figure~\ref{fig:bridges}). The result of merging
the Uberon core ontology with the species-specific ontologies along with the
corresponding bridges is a product called \emph{composite-metazoan} (hereafter
CM), a single, consistent cross-species ontology. The Cell Ontology (CL), an
ontology of cell types, follows the same approach~\cite{diehl2016}, and is
automatically included in Uberon's CM product.

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figs/bridges}
  \caption{Uberon and the Drosophila Anatomy Ontology (FBbt) maintain
  their own hierarchy of terms separately. Simply merging the two
  ontologies together \textbf{(A)} would result in a hierarchy that is
  apparently unified but is, in effect, split in two independent
  branches that are solely connected by BFO's \term{independent
  continuant} root term, preventing any meaningful use. \textbf{(B)}
  Bridging axioms (dotted red arrows) between Uberon and FBbt allow to
  connect \species{Drosophila}-specific terms to their taxon-neutral
  counterparts, thereby crossing the gap between the two branches.}
  \label{fig:bridges}
\end{figure}

To implement its strategy, Uberon maintains a set of cross-ontology
mappings, which are used to determine where the terms from the
species-specific ontologies should be placed in the overall Uberon
hierarchy. Across ontologies of the OBO Foundry, the typical method to
represent and maintain mappings is to use cross-references (or,
informally, ``xrefs''). It’s a very simple method where, to map a term
$T_A$ in an ontology A to a term $T_B$ in a foreign ontology B, $T_A$ is
annotated with a \property{oboInOwl:hasDbXref} property, whose value is
the short identifier (``CURIE'') of $T_B$. As part of the Uberon release
pipeline, a Perl script automatically extracts those cross-references
and generates the bridge files needed to integrate the species-specific
ontologies.

While simple, the cross-reference method has several limitations. (i) It
does not allow to record any metadata about the mapping, such as: who
asserted that the two terms should be mapped? On what basis? When was
the mapping reviewed? (ii) The method does not allow for any nuance:
either two terms are mapped or they are not, without room to express
more subtle relations. (iii) Cross-references have no clear and
universally agreed upon meaning. When two terms are mapped, the meaning
of the mapping is left unspecified, and typically has to be inferred
from the ontologies that are being mapped (for example, a mapping
between the species-neutral Uberon and the \species{Drosophila}-specific
FBbt can be inferred to be a cross-species mapping). This is made worse
by the fact that cross-references in OBO ontologies are used for many
different things beyond just mappings; for example, when used to
annotate a term definition, they typically attribute a source for the
definition, akin to a citation in an academic paper.

\section{New approach to cross-ontology mappings}

We have devised a new approach to maintain and use cross-ontology
mappings in Uberon, centered around three axes: (i) the use of a new
format to represent the mappings, (ii) the creation of dedicated mapping
relations, and (iii) the development of new tools to manipulate the
mappings and, in particular, derive OWL axioms from them.

\subsection{The SSSOM standard}

The Simple Standard for Sharing Ontological Mappings (SSSOM) is a recently
devised standard specifically intended to facilitate the exchange of semantic
mappings~\cite{matentzoglu2022b}. It allows treating mappings as first-class
data entities and to attach to them a range of metadata such as provenance and
licensing information. The standard defines a common data model to represent
mappings as well as two distinct serialisation formats to store and transport
instances of the model: a JSON-based format and a TSV-based format. The TSV
format has been specifically designed to be easily manipulatable both by
standard spreadsheet software (for editing by curators) and by common science
libraries and tools (for exploitation by data scientists and engineers). At
least two independent implementations of the standard are available: SSSOM-Py
and SSSOM-Java, for the Python and Java programming languages, respectively. In
addition, several ontology-related tools have started to add direct support for
the format, such as the Ontology Access Kit~\cite{mungall2023} and the Ontology
Development Kit~\cite{matentzoglu2022a}. The community resource Biomappings
also provides its mappings in SSSOM, among other formats~\cite{hoyt2023}.

\subsection{New mapping relations for cross-species mappings}

One of the benefits brought by the SSSOM standard is the possibility to use
highly specific mapping predicates to express precisely what is the intended
meaning of a mapping. While many mapping sets don’t actually need this, and can
use common mapping predicates such as those from the SKOS
vocabulary~\cite{bechhofer2009} (\property{skos:exactMatch},
\property{skos:narrowMatch}, etc.), cross-species mappings are an example of an
application where the common predicates are not satisfying. For example, let us
consider a mapping between the FBbt term for neuron (\curie{FBbt:00005106},
representing a neuron in a fruit fly) and the corresponding term in the Cell
Ontology (\curie{CL:0000540}, representing a neuron in any species): clearly it
would not be correct to state that a fly neuron is the same concept as a neuron
in any other species --~the two terms are not interchangeable, so
\property{skos:exactMatch} is not a suitable mapping predicate.  A slightly
more correct predicate could be \property{skos:narrowMatch}, because is it true
that a fly neuron is a narrower concept than a species-neutral neuron; but a
sensory neuron (\curie{CL:0000101}) is also a narrower concept than a generic
neuron, yet the relation between \term{sensory neuron} and \term{neuron} is of
a different nature than the relation between \term{fly neuron} and
\term{(species-neutral) neuron}.

Overall we believe that cross-species mappings are \emph{sui generis}
mappings and that they warrant the use of dedicated mapping predicates
to reflect that fact. We therefore expanded the SEMAPV vocabulary,
specifically established for use with SSSOM~\cite{matentzoglu2022}, with
four new mapping predicates (Figure~\ref{fig:relations}) that mirrors
existing SKOS predicates but are specifically intended for cases where
the subject and object of a mapping belong to different taxonomic
groupings.

\begin{figure}
  \centering
  \includegraphics[width=.8\linewidth]{figs/relations}
  \caption{Excerpt of the semantic mapping vocabulary (SEMAPV), with the
  new mapping relations (yellow) introduced to specifically represent
  cross-species mappings.}
  \label{fig:relations}
\end{figure}

\subsection{Deriving OWL bridges from SSSOM}

Even with fast SSSOM adoption, it is likely that many tools will remain
unable to directly exploit SSSOM sets in the foreseeable future.
Therefore, when mappings are stored in the SSSOM format, we need to be
able to convert them into OWL axioms that can be used by any ontology
manipulation tool. To that effect, we have developed a plugin for ROBOT,
the standard tool used to manipulate OBO ontologies~\cite{jackson2019}.
The plugin is built on top of the Java implementation of the SSSOM
standard and provides a \emph{sssom:inject} command that takes a SSSOM
mapping set as input, derives OWL axioms from the mappings, and injects
them into an ontology.

Rather than hardcoding the logic for deriving the axioms --~which would
have been efficient but would have made the logic harder to modify and
re-use~--, we designed a small domain-specific language (DSL), named
SSSOM/T-OWL (``SSSOM/Transform-to-OWL''), that allows users to describe
which mappings in a set should be transformed into OWL axioms and what
kind of axioms should be produced.

The SSSOM/T-OWL language is organised around the single concept of
``transformation rules''. A rule is made of two elements
(Table~\ref{tab:sssomt}): a filter (or selector) to decide whether the
rule should apply to a given mapping; and either a preprocessor, to
modify a mapping on the fly, or a generator, to produce a OWL axiom from
the mapping.

\begin{table}
  \caption{SSSOM/T-OWL filters, preprocessors, and generators}
  \label{tab:sssomt}
  \begin{tabular}{ll}
    \toprule
    SSSOM/T-OWL element & Examples\\
    \midrule
    Atomic filter & \verb|predicate==skos:exactMatch|\\
    Negated filter & \verb|!subject==UBERON:*|\\
    Intersection filter & \verb|subject==UBERON:* && predicate==skos:exactMatch|\\
    Union filter & \verb+subject==UBERON:* || subject==CL:*+\\
    \midrule
    Inversion preprocessor & \verb|-> invert()|\\
    Exclusion preprocessor & \verb|-> stop()|\\
    Assignation preprocessor & \verb|-> assign("subject_source", "uberon.owl")|\\
    Replacement preprocessor & \verb|-> replace("mapping_tool", "pattern", "replacement")|\\
    \midrule
    Generic axiom generator & \verb|-> create_axiom("%subject_id EquivalentTo: %object_id")|\\
    Subject annotation generator & \verb|-> annotate_subject(oboInOwl:hasDbXref, "%object_curie")|\\
    Object annotation generator & \verb|-> annotate_object(oboInOwl:hasDbXref, "%subject_curie")|\\
    \bottomrule
  \end{tabular}
\end{table}

Filters allow to compare the value of any of the SSSOM metadata slots
with a target value and to discard any mapping with a different value. A
basic form of wildcard matching is supported, where a filter can select
mappings with a value that starts with the same prefix as the target
value. Individual filters can be combined to select mappings based on
more than just one metadata slot; they can also be negated, so that a
filter that would select a given set of mappings would, after negation,
discard that set and instead select the complementary set of mappings.

Preprocessors allow modifying a mapping before it is used by subsequent
rules. Possible modifications include inverting a mapping (the subject
becomes the object and vice-versa) and changing some of the mapping's
metadata. A preprocessor can also be used to completely remove a mapping
from the set, so that no further rule will be applied to this mapping.

Finally, the generators take the selected mapping and turn it into an axiom.
The most important generator function is \function{create\_axiom}, which can
produce an arbitrary axiom described by an expression in Manchester
syntax~\cite{patel-schneider2012}. Other generators allow to annotate classes
of the ontology with any of the available mappings metadata.

Of note, the current implementation of the SSSOM/T-OWL language clearly
separates the OWL generators from the filters and the preprocessing
functions. This makes it easy to create other domain-specific languages
similar to SSSOM/T-OWL, using the same overall syntax, the same filters
and the same preprocessors, but that generate other objects than OWL
axioms.

\section{Application of the approach to the Uberon pipeline}

With the components described in the previous section in place, we
overhauled the pipeline that builds Uberon's \emph{composite-metazoan}
product (Figure~\ref{fig:pipeline}).

\begin{figure}
  \centering
  \includegraphics[width=.4\linewidth]{figs/pipeline}
  \caption{Uberon's new pipeline. A Single SSSOM set, encompassing
  mappings with all target species, is created by extracting
  old-style cross-references from both Uberon and several other
  species-specific anatomy ontologies --~except FlyBase's FBbt, which
  already provides a ready-to-use SSSOM set (1). The resulting set is
  processed by the Uberon SSSOM/T-OWL ruleset (2) to create the bridge
  files, which are then merged along with the source ontologies to
  create the \emph{collected-metazoan} ontology (3). The final
  \emph{composite-metazoan} is created (4) by pruning redundant terms
  using a custom ROBOT plugin.}
  \label{fig:pipeline}
\end{figure}

\subsection{Collecting the mappings}

Rather than migrating all of Uberon mappings to SSSOM at once, we have
adopted a phased approach, in which we have used the mappings between
Uberon and FBbt as a test bed in a preliminary phase. Uberon/FBbt
mappings were initially maintained as cross-references within Uberon, as
for all the other Uberon mappings. We migrated them to a SSSOM mapping
set that we now maintain on the FBbt side: existing FBbt
cross-references in Uberon were extracted from the ontology and
converted into SSSOM mappings using the newly minted
\property{semapv:crossSpeciesExactMatch} predicate (all existing
cross-references were for exact mappings, since that was the only kind
of mapping allowed by the cross-reference system), that are now
published alongside the FBbt ontology itself.

Mappings between Uberon and the other species-specific anatomy
ontologies beyond FBbt are still currently maintained as
cross-references either within Uberon or within the species-specific
ontologies. They are extracted and converted to SSSOM dynamically as
part of Uberon's release pipeline. In a second phase, which we have not
started yet, we will migrate those mappings to SSSOM definitely, as was
done for the FBbt mappings.

\subsection{Applying the transformation rules}

Regardless of their origin, all mappings between Uberon, CL, and the
species-specific ontologies are then processed by a single large
SSSOM/T-OWL ruleset that contains all the required logic to generate the
bridging axioms. A typical rule in this ruleset is the following:

\begin{lstlisting}
(subject==UBERON:* || subject==CL:*) && object==FBbt:*
  && predicate==semapv:crossSpeciesExactMatch
  -> create_axiom('%object_id EquivalentTo: %subject_id and
                   (BFO:0000050 some NCBITaxon:7227)');
\end{lstlisting}

That rule can be literally understood as: a mapping between a Uberon (or
CL) term and a FBbt term and that uses the
\property{semapv:crossSpeciesExactMatch} predicate must be transformed
into an equivalence axiom that states that the FBbt term is equivalent
to the intersection of the Uberon (or CL) term and an existential
restriction over \curie{BFO:0000050} (\term{part of}) and
\curie{NCBITaxon:7227} (\term{Drosophila melanogaster}).

All axioms produced by the SSSOM/T-OWL ruleset are dispatched into one
different bridge file for each foreign ontology.

\subsection{Building the composite-metazoan ontology}

Once the bridge files have been generated, an intermediate ontology
(\emph{collected-metazoan}) is built by merging together: Uberon itself,
the Cell Ontology, all the species-specific ontologies, and their
corresponding bridges. Finally, \emph{composite-metazoan} itself is
derived from this intermediate product by applying some custom logic
(implemented in a small, dedicated ROBOT plugin) to ``prune'' redundant
terms and replace them by equivalent anonymous class expressions.

The details of this last operation are beyond the scope of this paper,
but as an example, let us consider the FBbt term \term{ovary}
(\curie{FBbt:00004865}): it is mapped to the Uberon term \term{ovary}
(\curie{UBERON:0000992}), so \emph{collected-metazoan} contains the
following axiom:

\begin{lstlisting}
FBbt:00004865 EquivalentTo: UBERON:0000992 and
                            (BFO:0000050 some NCBITaxon:7227)
\end{lstlisting}

The ``pruning'' operation leading to \emph{composite-metazoan} consists
in replacing all occurrences of \curie{FBbt:00004865} by the anonymous
expression it is equivalent to. So the following axiom, which states
that the \term{oviduct} (\curie{FBbt:00004911}) is \term{continuous
with} (\curie{RO:0002150}) the fly ovary:

\begin{lstlisting}
FBbt:00004911 SubClassOf: RO:0002150 some FBbt:000004865
\end{lstlisting}

gets rewritten as:

\begin{lstlisting}
FBbt:00004911 SubClassOf: RO:0002150 some (UBERON:0000992 and
                          (BFO:0000050 some NCBITaxon:7227))
\end{lstlisting}

\section{Conclusion}

While full transition to SSSOM is not achieved yet, we now have the
needed infrastructure in place to complete this migration, and we have a
more robust and more flexible pipeline for creating the bridges needed
for \emph{composite-metazoan}. The logic for deriving the bridges is
expressed using a simple DSL that most editors should be able to master
quickly, should the need arise to modify said logic (for example to add
a new bridge). While the precise SSSOM/T-OWL rules we use are obviously
specific to the needs of Uberon, we believe the general approach of
maintaining mappings as SSSOM sets and transforming them into OWL
bridges using a simple DSL could be re-used by any project that requires
integrating several ontologies together.

\begin{acknowledgments}
This work was supported by grant BB/T014008 from the UK Biotechnology
and Biological Sciences Research Council (BBSRC) and the US National
Science Foundation Directorate of Biological Sciences (NSF/BIO).
\end{acknowledgments}

\bibliography{sssomt}

\appendix

\section{Online Resources}

The SSSOM-Java project, which implements both the SSSOM standard and the
SSSOM/T-OWL language described in this paper, is hosted on
\href{https://github.com/gouttegd/sssom-java/}{GitHub}.

\end{document}

### 2024-03-31 sssomt.tex
Suggested corrections from N. Harris.

@@
 
-One of the pre-requisites for the exploitation of datasets from
-high-throughput experimental methods, such as single-cell RNA sequencing
-datasets, beyond their lab of origin is the ability to annotate samples
-and results with consistent terms. That is, we need controlled
+One of the prerequisites for the reuse and renalysis of datasets from
+high-throughput experimental methods --~such as single-cell RNA sequencing
+datasets~-- beyond their lab of origin is the annotation of samples
+and results with standardised terms. That is, we need controlled
 vocabularies to properly record information such as the exact

### 2024-04-02 sssomt.tex
More corrections by N. Harris.

@@
 One of the prerequisites for the reuse and renalysis of datasets from
-high-throughput experimental methods --~such as single-cell RNA sequencing
-datasets~-- beyond their lab of origin is the annotation of samples
-and results with standardised terms. That is, we need controlled
-vocabularies to properly record information such as the exact
-techniques used, the organs or tissues from which the experimental
-samples were taken, or the cell types that were identified. Several
-ontologies have been developed for this kind of purpose, such as the
-Experimental Factor Ontology (EFO) to annotate experimental
-methods~\cite{malone2010}, the Drosophila Anatomy Ontology (FBbt) to
-annotate anatomical structures and cell types in
-\species{D.~melanogaster}~\cite{costa2013}, the Worm Anatomy Ontology
-(WBbt) to do the same in \species{C.~elegans}~\cite{lee2003}, etc. Most
-of those ontologies are grouped under the umbrella of the OBO
-Foundry~\cite{jackson2021}.
+high-throughput experimental methods --~such as single-cell RNA
+sequencing datasets~-- beyond their lab of origin is the annotation of
+samples and results with standardised terms. That is, we need controlled
+vocabularies to properly record information such as the exact techniques
+used, the organs or tissues from which the experimental samples were
+taken, or the cell types that were identified. Several ontologies have
+been developed for this kind of purpose, such as the Experimental Factor
+Ontology (EFO) to annotate experimental methods~\cite{malone2010}, the
+Drosophila Anatomy Ontology (FBbt) to annotate anatomical structures and
+cell types in \species{D.~melanogaster}~\cite{costa2013}, the Worm
+Anatomy Ontology (WBbt) to do the same in
+\species{C.~elegans}~\cite{lee2003}, etc. Most of those ontologies are
+grouped under the umbrella of the OBO Foundry~\cite{jackson2021}.
 
@@
 
-Building a single cross-species anatomy ontology from the ground up, with terms
-suitable for all the model organisms, would be a gigantic task, requiring lots
-of species-specific expertise that would be hard to gather in a single project.
-Instead, a more realistic approach was adopted by the Uberon
-project~\cite{mungall2012a} which, rather than attempting to describe the
-anatomy of every model organism, leverages the existing species-specific
-anatomy ontologies. Uberon aims to provide a ``core'' anatomy ontology that
-describes anatomical structures in a species-neutral way, along with
-``bridges'' that allow to integrate the more precise terms from the
-species-specific ontologies (Figure~\ref{fig:bridges}). The result of merging
-the Uberon core ontology with the species-specific ontologies along with the
-corresponding bridges is a product called \emph{composite-metazoan} (hereafter
-CM), a single, consistent cross-species ontology. The Cell Ontology (CL), an
-ontology of cell types, follows the same approach~\cite{diehl2016}, and is
-automatically included in Uberon's CM product.
+Building a single cross-species anatomy ontology from the ground up,
+with terms suitable for all the model organisms, would be a gigantic
+task, requiring lots of species-specific expertise that would be hard to
+gather in a single project.  Instead, a more practical approach was
+adopted by the Uberon project~\cite{mungall2012a} which, rather than
+attempting to describe the anatomy of every model organism, leverages
+the existing species-specific anatomy ontologies. Uberon aims to provide
+a ``core'' anatomy ontology that describes anatomical structures in a
+species-neutral way, along with ``bridges'' that allow to integrate the
+more precise terms from the species-specific ontologies
+(Figure~\ref{fig:bridges}). The result of merging the Uberon core
+ontology with the species-specific ontologies along with the
+corresponding bridges is a product called \emph{composite-metazoan}
+(hereafter CM), a single, consistent cross-species ontology. The Cell
+Ontology (CL), an ontology of cell types, follows the same
+approach~\cite{diehl2016}, and is automatically included in Uberon's CM
+product.
 
@@
 
-To implement its strategy, Uberon maintains a set of cross-ontology
+To implement this strategy, Uberon maintains a set of cross-ontology
 mappings, which are used to determine where the terms from the
@@
 more subtle relations. (iii) Cross-references have no clear and
-universally agreed upon meaning. When two terms are mapped, the meaning
-of the mapping is left unspecified, and typically has to be inferred
-from the ontologies that are being mapped (for example, a mapping
-between the species-neutral Uberon and the \species{Drosophila}-specific
-FBbt can be inferred to be a cross-species mapping). This is made worse
-by the fact that cross-references in OBO ontologies are used for many
-different things beyond just mappings; for example, when used to
-annotate a term definition, they typically attribute a source for the
-definition, akin to a citation in an academic paper.
+universally agreed upon meaning. When two terms are mapped to each
+other, the meaning of the mapping is left unspecified, and typically has
+to be inferred from the ontologies that are being mapped (for example, a
+mapping between the species-neutral Uberon and the
+\species{Drosophila}-specific FBbt can be inferred to be a cross-species
+mapping). This is made worse by the fact that cross-references in OBO
+ontologies are used for many different things beyond just mappings; for
+example, when used to annotate a term definition, cross-references
+typically attribute a source for the definition, akin to a citation in
+an academic paper.
 
@@
 We have devised a new approach to maintain and use cross-ontology
-mappings in Uberon, centered around three axes: (i) the use of a new
+mappings in Uberon, centred around three axes: (i) the use of a new
 format to represent the mappings, (ii) the creation of dedicated mapping
@@
 
-The Simple Standard for Sharing Ontological Mappings (SSSOM) is a recently
-devised standard specifically intended to facilitate the exchange of semantic
-mappings~\cite{matentzoglu2022b}. It allows treating mappings as first-class
-data entities and to attach to them a range of metadata such as provenance and
-licensing information. The standard defines a common data model to represent
-mappings as well as two distinct serialisation formats to store and transport
-instances of the model: a JSON-based format and a TSV-based format. The TSV
-format has been specifically designed to be easily manipulatable both by
-standard spreadsheet software (for editing by curators) and by common science
-libraries and tools (for exploitation by data scientists and engineers). At
-least two independent implementations of the standard are available: SSSOM-Py
-and SSSOM-Java, for the Python and Java programming languages, respectively. In
-addition, several ontology-related tools have started to add direct support for
-the format, such as the Ontology Access Kit~\cite{mungall2023} and the Ontology
-Development Kit~\cite{matentzoglu2022a}. The community resource Biomappings
-also provides its mappings in SSSOM, among other formats~\cite{hoyt2023}.
+The Simple Standard for Sharing Ontological Mappings (SSSOM) is a
+recently devised standard specifically intended to facilitate the
+exchange of semantic mappings~\cite{matentzoglu2022b}. It allows
+mappings to be treated as first-class data entities and to attach to
+them a range of metadata such as provenance and licensing information.
+The standard defines a common data model to represent mappings as well
+as two distinct serialisation formats to store and transport instances
+of the model: a JSON-based format and a TSV-based format. The TSV format
+has been specifically designed to be easily manipulatable both by
+standard spreadsheet software (for editing by curators) and by common
+science libraries and tools (for use by data scientists and engineers).
+At least two independent implementations of the standard are available:
+SSSOM-Py and SSSOM-Java, for the Python and Java programming languages,
+respectively. In addition, several ontology-related tools have started
+to add direct support for the format, such as the Ontology Access
+Kit~\cite{mungall2023} and the Ontology Development
+Kit~\cite{matentzoglu2022a}. The community resource Biomappings also
+provides its mappings in SSSOM, among other formats~\cite{hoyt2023}.
 
@@
 
-One of the benefits brought by the SSSOM standard is the possibility to use
-highly specific mapping predicates to express precisely what is the intended
-meaning of a mapping. While many mapping sets don’t actually need this, and can
-use common mapping predicates such as those from the SKOS
+One of the benefits brought by the SSSOM standard is the possibility to
+use highly specific mapping predicates to express precisely the intended
+meaning of a mapping. While many mapping sets don’t actually need this,
+and can use common mapping predicates such as those from the SKOS
 vocabulary~\cite{bechhofer2009} (\property{skos:exactMatch},
-\property{skos:narrowMatch}, etc.), cross-species mappings are an example of an
-application where the common predicates are not satisfying. For example, let us
-consider a mapping between the FBbt term for neuron (\curie{FBbt:00005106},
-representing a neuron in a fruit fly) and the corresponding term in the Cell
-Ontology (\curie{CL:0000540}, representing a neuron in any species): clearly it
-would not be correct to state that a fly neuron is the same concept as a neuron
-in any other species --~the two terms are not interchangeable, so
-\property{skos:exactMatch} is not a suitable mapping predicate.  A slightly
-more correct predicate could be \property{skos:narrowMatch}, because is it true
-that a fly neuron is a narrower concept than a species-neutral neuron; but a
-sensory neuron (\curie{CL:0000101}) is also a narrower concept than a generic
-neuron, yet the relation between \term{sensory neuron} and \term{neuron} is of
-a different nature than the relation between \term{fly neuron} and
-\term{(species-neutral) neuron}.
+\property{skos:narrowMatch}, etc.), cross-species mappings are an
+example of an application where the common predicates are not
+sufficient. For example, let us consider a mapping between the FBbt term
+for neuron (\curie{FBbt:00005106}, representing a neuron in a fruit fly)
+and the corresponding term in the Cell Ontology (\curie{CL:0000540},
+representing a neuron in any species): clearly it would not be correct
+to state that a fly neuron is the same concept as a neuron in any other
+species --~the two terms are not interchangeable, so
+\property{skos:exactMatch} is not a suitable mapping predicate.  A
+slightly more correct predicate could be \property{skos:narrowMatch},
+because is it true that a fly neuron is a narrower concept than a
+species-neutral neuron; but a sensory neuron (\curie{CL:0000101}) is
+also a narrower concept than a generic neuron, yet the relation between
+\term{sensory neuron} and \term{neuron} is of a different nature than
+the relation between \term{fly neuron} and \term{(species-neutral)
+neuron}.
 
@@
 specifically established for use with SSSOM~\cite{matentzoglu2022}, with
-four new mapping predicates (Figure~\ref{fig:relations}) that mirrors
+four new mapping predicates (Figure~\ref{fig:relations}) that mirror
 existing SKOS predicates but are specifically intended for cases where
@@
 
-Even with fast SSSOM adoption, it is likely that many tools will remain
-unable to directly exploit SSSOM sets in the foreseeable future.
-Therefore, when mappings are stored in the SSSOM format, we need to be
-able to convert them into OWL axioms that can be used by any ontology
-manipulation tool. To that effect, we have developed a plugin for ROBOT,
-the standard tool used to manipulate OBO ontologies~\cite{jackson2019}.
-The plugin is built on top of the Java implementation of the SSSOM
-standard and provides a \emph{sssom:inject} command that takes a SSSOM
-mapping set as input, derives OWL axioms from the mappings, and injects
-them into an ontology.
+Even as SSSOM adoption spreads rapidly, it is likely that many tools
+will remain unable to directly exploit SSSOM sets in the foreseeable
+future.  Therefore, when mappings are stored in the SSSOM format, we
+need to be able to convert them into OWL axioms that can be used by any
+ontology manipulation tool. To that effect, we have developed a SSSOM
+plugin for ROBOT, the standard tool used to manipulate OBO
+ontologies~\cite{jackson2019}.  The plugin is built on top of the Java
+implementation of the SSSOM standard and provides a \emph{sssom:inject}
+command that takes a SSSOM mapping set as input, derives OWL axioms from
+the mappings, and injects them into an ontology.
 
@@
 have been efficient but would have made the logic harder to modify and
-re-use~--, we designed a small domain-specific language (DSL), named
+re-use~-- we designed a small domain-specific language (DSL), named
 SSSOM/T-OWL (``SSSOM/Transform-to-OWL''), that allows users to describe
@@
 
-Filters allow to compare the value of any of the SSSOM metadata slots
-with a target value and to discard any mapping with a different value. A
-basic form of wildcard matching is supported, where a filter can select
-mappings with a value that starts with the same prefix as the target
-value. Individual filters can be combined to select mappings based on
-more than just one metadata slot; they can also be negated, so that a
-filter that would select a given set of mappings would, after negation,
-discard that set and instead select the complementary set of mappings.
+Filters allow the value of any of the SSSOM metadata slots to be
+compared with a target value and to discard any mapping with a different
+value. A basic form of wildcard matching is supported, where a filter
+can select mappings with a value that starts with the same prefix as the
+target value. Individual filters can be combined to select mappings
+based on more than just one metadata slot; they can also be negated, so
+that a filter that would select a given set of mappings would, after
+negation, instead select the complementary set of mappings.
 
-Preprocessors allow modifying a mapping before it is used by subsequent
-rules. Possible modifications include inverting a mapping (the subject
-becomes the object and vice-versa) and changing some of the mapping's
-metadata. A preprocessor can also be used to completely remove a mapping
-from the set, so that no further rule will be applied to this mapping.
+Preprocessors allow a mapping to be modified before it is used by
+subsequent rules. Possible modifications include inverting a mapping
+(the subject becomes the object and vice-versa) and changing some of the
+mapping's metadata. A preprocessor can also be used to completely remove
+a mapping from the set, so that no further rule will be applied to this
+mapping.
 
-Finally, the generators take the selected mapping and turn it into an axiom.
-The most important generator function is \function{create\_axiom}, which can
-produce an arbitrary axiom described by an expression in Manchester
-syntax~\cite{patel-schneider2012}. Other generators allow to annotate classes
-of the ontology with any of the available mappings metadata.
+Finally, the generators take the selected mapping and turn it into an
+axiom.  The most important generator function is
+\function{create\_axiom}, which can produce an arbitrary axiom described
+by an expression in Manchester syntax~\cite{patel-schneider2012}. Other
+generators allow classes of the ontology to be annotated with any of the
+available mappings metadata.
 
-Of note, the current implementation of the SSSOM/T-OWL language clearly
+Of note, the current implementation of the SSSOM/T-OWL language cleanly
 separates the OWL generators from the filters and the preprocessing
@@
 similar to SSSOM/T-OWL, using the same overall syntax, the same filters
-and the same preprocessors, but that generate other objects than OWL
+and the same preprocessors, but that generate objects other than OWL
 axioms.
@@
 part of Uberon's release pipeline. In a second phase, which we have not
-started yet, we will migrate those mappings to SSSOM definitely, as was
-done for the FBbt mappings.
+started yet, we will migrate those mappings fully to SSSOM, as was done
+for the FBbt mappings.
 
@@
 Regardless of their origin, all mappings between Uberon, CL, and the
-species-specific ontologies are then processed by a single large
-SSSOM/T-OWL ruleset that contains all the required logic to generate the
-bridging axioms. A typical rule in this ruleset is the following:
+species-specific ontologies are processed by a single large SSSOM/T-OWL
+ruleset that contains all the required logic to generate the bridging
+axioms. A typical rule in this ruleset is the following:
 
@@
 
-All axioms produced by the SSSOM/T-OWL ruleset are dispatched into one
-different bridge file for each foreign ontology.
+All axioms produced by the SSSOM/T-OWL ruleset are saved into a separate
+bridge file for each foreign ontology.
 
@@
 Once the bridge files have been generated, an intermediate ontology
-(\emph{collected-metazoan}) is built by merging together: Uberon itself,
+(\emph{collected-metazoan}) is built by merging together Uberon itself,
 the Cell Ontology, all the species-specific ontologies, and their
@@
 (implemented in a small, dedicated ROBOT plugin) to ``prune'' redundant
-terms and replace them by equivalent anonymous class expressions.
+terms and replace them with equivalent anonymous class expressions.
 
@@
 The ``pruning'' operation leading to \emph{composite-metazoan} consists
-in replacing all occurrences of \curie{FBbt:00004865} by the anonymous
+of replacing all occurrences of \curie{FBbt:00004865} by the anonymous
 expression it is equivalent to. So the following axiom, which states
@@
 
-While full transition to SSSOM is not achieved yet, we now have the
-needed infrastructure in place to complete this migration, and we have a
-more robust and more flexible pipeline for creating the bridges needed
-for \emph{composite-metazoan}. The logic for deriving the bridges is
-expressed using a simple DSL that most editors should be able to master
-quickly, should the need arise to modify said logic (for example to add
-a new bridge). While the precise SSSOM/T-OWL rules we use are obviously
-specific to the needs of Uberon, we believe the general approach of
-maintaining mappings as SSSOM sets and transforming them into OWL
-bridges using a simple DSL could be re-used by any project that requires
-integrating several ontologies together.
+To construct a consistent and maintainable cross-species anatomy
+ontology, we have updated the integration mechanism between Uberon, CL,
+and the species-centric ontologies by using SSSOM to manage the mappings
+betweeen the ontologies. While full transition to SSSOM is not yet
+complete, we now have the needed infrastructure in place to complete
+this migration, and we have a more robust and more flexible pipeline for
+creating the bridges needed for \emph{composite-metazoan}. The logic for
+deriving the bridges is expressed using a simple DSL (SSSOM/T-OWL) that
+most ontology editors should be able to master quickly, should the need
+arise to modify said logic (for example to add a new bridge). While the
+precise SSSOM/T-OWL rules we use are obviously specific to the needs of
+Uberon, we believe the general approach of maintaining mappings as SSSOM
+sets and transforming them into OWL bridges using a simple DSL could be
+used by any project that requires integrating several ontologies
+together.
 

### 2024-04-02 sssomt.tex
Typo fix.

@@
 
-One of the prerequisites for the reuse and renalysis of datasets from
+One of the prerequisites for the reuse and reanalysis of datasets from
 high-throughput experimental methods --~such as single-cell RNA

### 2024-05-23 sssomt.tex
Add PURLs to Uberon artefacts.
None of these artefacts are new and therefore do not really belong here
IMO (that gives the impression they are a novelty of this paper, while
they are not, at least not for the bridges and the Composite Metazoan
product; what is new is _how_ we generate them), but reviewer #4 asked
for it.

@@
 
+Uberon artefacts concerned by this project are available under the
+following permanent URLs:
+
+\begin{tabular}{ll}
+\hline
+Artefact & PURL\\
+\hline
+SSSOM mapping set & \url{http://purl.obolibrary.org/obo/uberon/uberon.sssom.tsv}\\
+Bridge to ontology \emph{X} & \url{http://purl.obolibrary.org/obo/uberon/bridge/uberon-bridge-to-X.owl}\\
+Composite Metazoan & \url{http://purl.obolibrary.org/obo/uberon/composite-metazoan.owl}\\
+\hline
+\end{tabular}
+
 \end{document}

### 2024-05-28 presentation.tex
Initial commit.

\usepackage{fontspec}
\usepackage{polyglossia}
\usepackage{tikz}

\setmainlanguage{english}
\usetheme{cambridge}

\graphicspath{{imgs/}{svgs/}}

% TikZ libraries and styles
\usetikzlibrary{positioning,shapes.geometric,graphs,graphdrawing,quotes,fit}
\usegdlibrary{layered,trees}
\colorlet{GenericItem}{yellow!20!white}
\colorlet{GenericHighlight}{red!20!white}
\tikzset{Term/.style={draw,fill=yellow!20!white,font=\tiny\bfseries}}
\tikzset{Relation/.style={draw=blue!40!white,thick,arrows=<-}}
\tikzset{RelationLabel/.style={shape=circle,fill=white,draw=black,thin,font=\tiny\bfseries,inner sep=1pt}}
\tikzset{File/.style={draw,fill=yellow!20!white,font=\tiny\ttfamily}}
\tikzset{Entity/.style={draw,fill=yellow!20!white,font=\scriptsize}}
\tikzset{Process/.style={draw,shape=ellipse,fill=blue!20!white,font=\scriptsize}}

\title{(Re-)bridging the anatomy ontologies with SSSOM}

\author{Damien~Goutte-Gattat\inst{1}}
\institute{\inst{1}FlyBase group, Department of Physiology, Development and Neuroscience}
\date{International Conference on Biomedical Ontologies 2024, July 2024}

\hypersetup{pdfpagemode=UseNone,
  pdftitle={},
  pdfauthor={Damien Goutte-Gattat}}
\AtBeginSection{
  \begin{frame}{Outline}{}
    \tableofcontents[currentsection]
  \end{frame}
}

\begin{document}

\begin{frame}
  \titlepage
\end{frame}

\section{Introduction}

\begin{frame}
  \frametitle{An approach to a multi-species anatomy ontology}

  \begin{columns}
    \begin{column}{.5\textwidth}
      \begin{tikzpicture}[nodes={shape=circle,draw,fill=GenericItem}]
        \node (core) {Uberon};
        \node (fbbt) at (150:2cm) {FBbt};
        \node (zfa)  at (90:2cm)  {ZFA};
        \node (xao)  at (30:2cm)  {XAO};
        \node (ma)   at (210:2cm) {MA};
        \node (wbbt) at (270:2cm) {WBbt};
        \node (tx1) at  (310:2cm) {...};
        \node (tx2) at  (330:2cm) {...};
        \node (tx3) at  (350:2cm) {...};
        \draw (core) -- (fbbt) (core) -- (zfa)  (core) -- (xao)
              (core) -- (ma)   (core) -- (wbbt) (core) -- (tx1)
              (core) -- (tx2)  (core) -- (tx3);
      \end{tikzpicture}
    \end{column}
    \begin{column}{.45\textwidth}
      Strategy:
      \begin{itemize}
        \item Provide a taxon-neutral core framework
        \item Integrate existing taxon-specific ontologies
      \end{itemize}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}
  \frametitle{Example: integration with FBbt}

  \only<1>{
    \begin{columns}[t]
      \begin{column}{.45\textwidth}
        Uberon hierarchy
        \begin{tikzpicture}
          \graph[layered layout,orient=down,nodes={Term},edge quotes mid,edges={Relation,nodes={RelationLabel}}] {
            anatomical entity <-
            anatomical structure <-
            circulatory organ <- {
              primary circulatory organ <- { heart, dorsal vessel heart },
              accessory circulatory organ <- lymph heart
            };
          };
        \end{tikzpicture}
      \end{column}
      \begin{column}{.45\textwidth}
        FBbt hierarchy
        \begin{tikzpicture}
          \graph[layered layout,orient=down,nodes={Term},edge quotes mid,edges={Relation,nodes={RelationLabel}}] {
            anatomical entity <-
            anatomical structure <-
            heart <-["P"] {
              aortic funnel, heart chamber, dorsal diaphragm
            };
          };
        \end{tikzpicture}
      \end{column}
    \end{columns}
  }
  \only<2>{
    \begin{columns}
      \begin{column}{.3\textwidth}
        Integrated hierarchy
      \end{column}
      \begin{column}{.6\textwidth}
        \begin{tikzpicture}
          \graph[layered layout,orient=down,nodes={Term},edge quotes mid,edges={Relation,nodes={RelationLabel}}] {
            anatomical entity <-
            anatomical structure <-
            circulatory organ <- {
              primary circulatory organ <- {
                heart,
                dorsal vessel heart <- "heart (drosophila)" <-["P"] {
                  aortic funnel, heart chamber, dorsal diaphragm
                }
              },
              accessory circulatory organ <- lymph heart
            };
          };
        \end{tikzpicture}
      \end{column}
    \end{columns}
  }
\end{frame}

\begin{frame}
  \frametitle{Integration requires cross-ontology mappings}

  \begin{block}{}
    Integrating the taxon-specific ontologies with the Uberon core framework
    requires to know where taxon-specific terms should be ``plugged'' in the
    core framework.

    E.g., we need to know that FBbt's \emph{heart} must be attached to Uberon's
    \emph{dorsal vessel heart}.
  \end{block}
\end{frame}

\section{Current system}

\begin{frame}
  \frametitle{Current solution: cross-references}

  \begin{block}{}\ttfamily
[Term]

id: UBERON:0015230

name: dorsal vessel heart

def: "The caudal, pulsatile region of the dorsal vessel of the arthropod circulatory system."

is\_a: UBERON:0007100 ! primary circulatory organ

xref: FBbt:00003154
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{From cross-references to bridge files}

  \begin{columns}
    \begin{column}{.3\textwidth}
      \begin{tikzpicture}
        \node [File] (uberon) at (0,0)     {uberon.obo};
        \node [File] (script) at (0,-1)    {make-bridge-ontology-from-xrefs.pl};
        \node [File] (bridge) at (0,-2)    {uberon-bridge-to-fbbt.obo};
        \node [File]          at (0, -2.5) {uberon-bridge-to-wbbt.obo};
        \node [File]          at (0, -3)   {uberon-bridge-to-ma.obo};
        \node [File]          at (0, -3.5) {uberon-bridge-to-....obo}; 
        \draw[->] (uberon) -- (script) (script) -- (bridge);
      \end{tikzpicture}
    \end{column}
    \begin{column}{.68\textwidth}
      \begin{block}{Bridge file}\ttfamily
[Term]

id: FBbt:00003154

property\_value: IAO:0000589 "heart (drosophila)" xsd:string

intersection\_of: UBERON:0015230

intersection\_of: part\_of NCBITaxon:7227
      \end{block}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}
  \frametitle{Limitations of the current system}

  Limitations of cross-references to represent mappings
  \begin{itemize}
    \item Overloading of the \emph{oboInOwl:hasDbXref} annotation
    \item No metadata
    \item Mappings are embedded within the ontology and must be extracted
  \end{itemize}

  \bigskip
  Limitations of the bridging pipeline
  \begin{itemize}
    \item Highly Uberon-specific -- cannot be reused outside of Uberon
    \item Dependent on the OBO format
  \end{itemize}
\end{frame}

\section{New approach}

\begin{frame}
  \frametitle{Principles}

  \begin{enumerate}
    \item Clear separation between cross-ontology \emph{mappings} (COM) and cross-ontology \emph{bridges} (COB)
      \begin{itemize}
        \item A COB is merely the OWL materialisation of a COM
      \end{itemize}
    \item Represent COM using a standard format intended for that purpose
      \begin{itemize}
        \item allowing for mapping-specific metadata
        \item allowing for meaningful mapping predicates
      \end{itemize}
    \item Make it easy to derive COB from COM
      \begin{itemize}
        \item without requiring Uberon-specific tooling
      \end{itemize}
  \end{enumerate}
\end{frame}

\begin{frame}
  \frametitle{The SSSOM standard}

  ``Simple Standard for Sharing Ontological Mappings''
  \begin{itemize}
    \item A data model to represent semantic mappings and associated metadata
    \item A simple TSV-based format to store and exchange them
  \end{itemize}

  \begin{block}{SSSOM/TSV format}\tiny\ttfamily
    \begin{tabular}{lllll}
      \hline
      subject\_id   & subject\_label                & predicate\_id                 & object\_id      & mapping\_justification\\
      \hline
      FBbt:00000001 & organism                      & semapv:crossSpeciesExactMatch & UBERON:0000468  & semapv:ManualMappingCuration\\
      FBbt:00000002 & tagma                         & semapv:crossSpeciesExactMatch & UBERON:6000002  & semapv:ManualMappingCuration\\
      ...           & ...                           & ...                           & ...             & ...\\
      FBbt:00003154 & heart                         & semapv:crossSpeciesExactMatch & UBERON:0015230  & semapv:ManualMappingCuration\\
      ...           & ...                           &...                           & ...             & ...\\
      \hline
    \end{tabular}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{New mapping predicates for cross-species mappings}

  Additions to the \emph{Semantic Mapping Vocabulary} (SEMAPV)
  \begin{center}
    \begin{tikzpicture}
      \graph[tree layout,nodes={Term},edges={Relation}] {
        "skos:mappingRelation" <- {
          "semapv:isomorphicMatch" <- {
            "skos:exactMatch",
            "semapv:crossSpeciesExactMatch" [fill=GenericHighlight]
          },
          "semapv:nonIsomorphicMatch" <- {
            "skos:broadMatch" <- "semapv:crossSpeciesBroadMatch" [fill=GenericHighlight],
            "skos:narrowMatch" <- "semapv:crossSpeciesNarrowMatch" [fill=GenericHighlight]
          }
        };
      };
    \end{tikzpicture}
  \end{center}
\end{frame}

\begin{frame}[t]
  \frametitle{From SSSOM sets to bridge files: SSSOM/T-OWL}

  A domain-specific language to transform SSSOM mappings into OWL axioms

  \only<1>{
    \begin{block}{Model of a mapping processing rule}
      \begin{center}
        \begin{tikzpicture}
          \node [text width=5cm,fill=GenericItem,draw] (filter) at (0,0)
                {\textbf{Filter}\\
                 Mapping $\rightarrow$ boolean\\
                 Decide whether the rule applies to a mapping};
          \node [text width=5cm,fill=GenericItem,draw] (preprocessor) at (6,1)
                {\textbf{Preprocessor}\\
                 Mapping $\rightarrow$ mapping\\
                 Modify a mapping on the fly};
          \node [text width=5cm,fill=GenericItem,draw] (generator) at (6,-1)
                {\textbf{Generator}\\
                 Mapping $\rightarrow$ OWL axiom\\
                 Derive an axiom from a mapping};
          \draw[->] (filter) -- (preprocessor);
          \draw[->] (filter) -- (generator);
        \end{tikzpicture}
      \end{center}
    \end{block}
  }
  \only<2>{
    \begin{block}{Inverting a mapping}\ttfamily\scriptsize
subject==UBERON:* -> invert();
    \end{block}

    \begin{block}{Generating a simple axiom}\ttfamily\scriptsize
predicate==skos:broadMatch -> create\_axiom("\%subject\_id SubClassOf: \%object\_id");
    \end{block}

    \begin{block}{Generating a more complex axiom}\ttfamily\scriptsize
subject==FBbt:*\\
\hspace{1cm} \&\& predicate==semapv:crossSpeciesExactMatch\\
\hspace{1cm} \&\& (mapping\_justification==semapv:ManualMappingCuration || confidence>=0.8)\\
->  create\_axiom("\%subject\_id EquivalentTo: \%object\_id and\\
\hspace{2cm} (BFO:0000050 some NCBITaxon:7227"));
    \end{block}
  }
\end{frame}

\begin{frame}
  \frametitle{Implementation in ROBOT commands}

  \begin{block}{sssom:xref-extract command}
    Convert existing mappings represented as cross-references into a SSSOM set

    \medskip
    {\ttfamily\scriptsize
robot sssom:xref-extract -i input.owl --mapping-file output.sssom.tsv
    }
  \end{block}

  \begin{block}{sssom:inject command}
    Apply SSSOM/T-OWL rules to a mapping set and inject the resulting axioms into an ontology

    \medskip
    {\ttfamily\scriptsize
robot sssom:inject -i input.owl --sssom mappings.sssom.tsv --ruleset bridge.rules
    }
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Integrating the species-specific ontologies}

  \begin{center}
    \begin{tikzpicture}[node distance=.5cm]
      \node [Entity] (uberon) {Uberon};
      \node [Entity,right=of uberon] (com) {Mapping sets};
      \node [Entity,right=of com] (ssao) {Taxon-specific ontologies};
      \node [Process,below=of com] (sssomt) {SSSOM/T-OWL};
      \node [Entity,below=of sssomt] (cob) {Bridge files};
      \node [Process,below=of cob] (merge) {OWL Merge};
      \node [Entity,below=of merge] (cm) {Multi-species ontology};
      \draw [->] (com) -- (sssomt) (sssomt) -- (cob);
      \draw [->] (cob) -- (merge) (merge) -- (cm);
      \draw [->] (uberon.south) |- (merge.west);
      \draw [->] (ssao.south) |- (merge.east);
    \end{tikzpicture}
  \end{center}
\end{frame}

\section{Conclusion}

\begin{frame}
  \frametitle{Outcome}

  \begin{block}{Cross-ontology mappings as first-class artefacts}
    \begin{itemize}
      \item With their own metadata
      \item Can be maintained separately
      \item Available as a separate release product in a standard format
    \end{itemize}
  \end{block}

  \begin{block}{Generic framework for SSSOM-based integration of multiple ontologies}
    \begin{itemize}
      \item SSSOM/T-OWL is completely independent of Uberon
      \item Only the ruleset is specific
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Example: FBbt/Uberon/CL integration test}

  A single ROBOT command to merge FBbt with Uberon, CL, and the mappings-derived
  bridging axioms, and check for possible inconsistencies that would reveal
  incorrect mappings:

  \begin{block}{}
\begin{verbatim}
robot merge -i fbbt.owl -i uberon.owl -i cl.owl \
      sssom:inject --sssom mappings.sssom.tsv   \
                   --ruleset bridging.rules     \
      reason -r ELK
\end{verbatim}
  \end{block}
\end{frame}

\end{document}

### 2024-05-28 presentation.tex
Rework "current system" and "conclusion".

@@
 \begin{frame}
-  \frametitle{Current solution: cross-references}
+  \frametitle{Maintaining mappings as cross-references}
 
@@
   \end{block}
+
+  Limitations:
+  \begin{itemize}
+    \item Overloading of the \emph{oboInOwl:hasDbXref} annotation
+    \item No metadata
+    \item Mappings are embedded within the ontology and must be extracted
+  \end{itemize}
 \end{frame}
@@
     \begin{column}{.68\textwidth}
-      \begin{block}{Bridge file}\ttfamily
+      \begin{block}{uberon-bridge-to-fbbt.obo}\ttfamily
 [Term]
@@
 id: FBbt:00003154
-
-property\_value: IAO:0000589 "heart (drosophila)" xsd:string
 
@@
       \end{block}
+
+      Can be used to integrate FBbt into Uberon
     \end{column}
   \end{columns}
+
+  \medskip
+  Limitations:
+  \begin{itemize}
+    \item Highly Uberon-specific code -- cannot be reused outside of Uberon
+    \item Dependent on the OBO format
+  \end{itemize}
 \end{frame}
@@
 \begin{frame}
-  \frametitle{Limitations of the current system}
+  \frametitle{Integration pipeline}
 
-  Limitations of cross-references to represent mappings
+  \begin{columns}
+    \begin{column}{.44\textwidth}
+      \begin{tikzpicture}[node distance=.5cm]
+        \node [Entity] (uberon) {Uberon};
+        \node [Process,below right=of uberon] (script) {Custom script};
+        \node [Entity,above right=of script,align=left] (ssao) {Taxon-specific\\
+                                                                ontologies};
+        \node [Entity,below=of script] (cob) {Bridge files};
+        \node [Process,below=of cob] (merge) {OWL merge};
+        \node [Entity,below=of merge] (cm) {Multi-species ontology};
+        \draw [->] (uberon.east) -| (script.north) (script.south) -- (cob.north);
+        \draw [->] (cob) -- (merge) (merge) -- (cm);
+        \draw [->] (uberon.south) |- (merge.west);
+        \draw [->] (ssao.south) |- (merge.east);
+      \end{tikzpicture}
+    \end{column}
+    \begin{column}{.55\textwidth}
+      Limitations:
       \begin{itemize}
-    \item Overloading of the \emph{oboInOwl:hasDbXref} annotation
-    \item No metadata
-    \item Mappings are embedded within the ontology and must be extracted
+        \item Lack of flexibility
+          \begin{itemize}
+            \item Cannot integrate foreign mappings
+            \item Cannot generate a single bridge file
           \end{itemize}
-
-  \bigskip
-  Limitations of the bridging pipeline
-  \begin{itemize}
-    \item Highly Uberon-specific -- cannot be reused outside of Uberon
-    \item Dependent on the OBO format
+        \item Mappings not available independently
+        \item Not generalizable to other ontologies
       \end{itemize}
+    \end{column}
+  \end{columns}
 \end{frame}
@@
   \begin{enumerate}
-    \item Clear separation between cross-ontology \emph{mappings} (COM) and cross-ontology \emph{bridges} (COB)
+    \item Clear separation between
       \begin{itemize}
-        \item A COB is merely the OWL materialisation of a COM
+        \item cross-ontology \emph{mappings} (COM)
+        \item cross-ontology \emph{bridges} (COB)
       \end{itemize}
@@
 \begin{frame}
-  \frametitle{Integrating the species-specific ontologies}
+  \frametitle{New SSSOM-based integration pipeline}
 
@@
       \node [Entity,right=of uberon] (com) {Mapping sets};
-      \node [Entity,right=of com] (ssao) {Taxon-specific ontologies};
+      \node [Entity,right=of com,align=left] (ssao) {Taxon-specific\\
+                                                     ontologies};
       \node [Process,below=of com] (sssomt) {SSSOM/T-OWL};
       \node [Entity,below=of sssomt] (cob) {Bridge files};
-      \node [Process,below=of cob] (merge) {OWL Merge};
+      \node [Process,below=of cob] (merge) {OWL merge};
       \node [Entity,below=of merge] (cm) {Multi-species ontology};
@@
 \begin{frame}
-  \frametitle{Outcome}
+  \frametitle{Benefits of the new approach}
 

### 2024-06-06 presentation.tex
Add speaker notes.

@@
   \end{columns}
+
+  \note{Anatomy ontologies are used in many aspects of biocuration, for
+  example to capture the organ of origin of a biological sample, the tissue
+  where a gene is expressed, or the anatomical location where a disease
+  manifests itself.
+
+  Many species-specific curation projects have, therefore, their own ontology
+  to describe the anatomy of their favourite organism. For example, at FlyBase
+  we have our own \emph{Drosophila} Anatomy Ontology (FBbt), ZFIN has its
+  Zebrafish Anatomy Ontology (ZFA), XenBase has its Xenopus Anatomy Ontology
+  (XAO), and so on.
+
+  But to enable \emph{cross-species} curation and analysis of data (especially
+  high-throughput data, such as scRNAseq), we need ideally a single,
+  multi-species ontology that covers all our model organisms. For the past 12
+  years or so, this has been provided by Uberon, the Uber Anatomy Ontology.
+
+  Rather than duplicating the work already done by pre-existing,
+  species-specific ontologies, Uberon's strategy is to provide a taxon-neutral
+  core framework, describing common anatomical structures in broadly generic
+  terms, and to plug the species-specific ontologies into that core.}
+
 \end{frame}
@@
   }
+
+  \note<1>{To illustrate this strategy, here is (on the left) a simplified
+  excerpt of Uberon's hierarchy around the concept of ``heart'', which in
+  Uberon parlance is called, more generically, the ``primary circulatory
+  organ'' (to make room for the \emph{accessory} circulatory organs that can
+  be found in some species).
+
+  Below ``primary circulatory organ'', Uberon defines both ``heart'', which is
+  intended to refer to the \emph{vertebrate} heart, and ``dorsal vessel
+  heart'', which refers to the heart of arthropods such as fruit flies. Uberon
+  itself does not describe the arthropod heart in any more details -- there
+  are no terms under ``dorsal vessel heart''. That term is in fact merely
+  intended as a ``branching point'', where more specific terms from
+  arthropod-specific ontologies can be rattached.
+
+  In the \emph{Drosophila} Anatomy Ontology, we have a single term ``heart''
+  to refer to the dorsal vessel heart (we don't need to call it ``dorsal
+  vessel heart'' since it is the only heart that exists in flies), and below
+  that term, we have a handful of terms that describe more precisely the
+  different parts that make up the fly heart.}
+
+  \note<2>{When FBbt is integrated into Uberon, its ``heart'' term is
+  automatically treated as a subclass of the more generic ``dorsal vessel
+  heart'', thereby creating a unified hierarchy where the
+  \emph{Drosophila}-specific terms are appropriately rattached to Uberon's
+  taxon-neutral core.
+
+  Using this strategy, Uberon is able to accurately and precisely describe the
+  anatomy of the most used model organisms, by benefitting from all the
+  existing species-specific ontologies without having to duplicate their
+  efforts.
+
+  This talk is about how this integration process works.}
+
 \end{frame}
@@
   \end{block}
+
+  \note{The key component of the integration process is the knowledge of which
+  species-specific terms, from each of the species-specific ontologies such as
+  FBbt, should be ``attached'' to which taxon-neutral term in Uberon's core
+  framework. For example, knowing that FBbt's ``heart'' is the fly-specific
+  term for what Uberon calls the ``dorsal vessel heart''.
+
+  This is the role of cross-ontology \emph{mappings} -- basically, a mapping
+  is a statement that one term from an ontology A is ``mapped'', or
+  ``related'', to another term in an ontology B.}
+
 \end{frame}
@@
   \end{itemize}
+
+  \note{In the current system, mappings are represented and maintained as
+  \emph{cross-references}, that is, as a \emph{oboInOwl:hasDbXref} annotation
+  that is carried by the taxon-neutral term in Uberon and that points to the
+  taxon-specific term in another ontology -- as shown here, where Uberon's
+  ``dorsal vessel heart'' term carries a ``xref'' to FBbt:00003154, which is
+  FBbt's term for the fly heart that we have seen before.
+
+  That's a pretty simple way of representing mappings, but it has several
+  drawbacks.
+
+  First, the \emph{oboInOwl:hasDbXref} annotation has been used for a lot of
+  things since its introduction, and no longer has a clear and unambiguous
+  meaning -- what it means depends on where it is used.
+
+  Second, all this representation allows us to say is that there is a mapping,
+  but we can't know anything else about said mapping. Why does that mapping
+  exist? Who asserted it? When? On what basis? This is simply not recorded.
+
+  Third, the mappings are not available separately from the entire ontology.
+  This makes reusing those mappings more difficult than it should be
+  (difficult to Find, Access and Reuse).}
+
 \end{frame}
@@
   \end{itemize}
+
+  \note{Whenever we make a new Uberon release, the mappings (stored as
+  cross-references within Uberon) are extracted by a specialised Perl script
+  to produce a set of \emph{bridge files} (one for each species-specific
+  ontology). The bridge files are a OBO materialisation of the mappings; they
+  contain minimalist frames that serve to rattach a species-specific term to a
+  core Uberon term, as in this example, which states that FBbt:00003154 (the
+  fly heart) is equivalent to a Uberon's ``dorsal vessel heart'' that is part
+  of the \emph{Drosophila melanogaster} taxon.
+
+  The main problem with this derivation is that the aforementioned script is
+  highly specific to Uberon. Most of its behaviours are hard-coded, and the
+  script is virtually impossible to re-use outside of Uberon. This means,
+  among other things, that third-party projects that might need those bridges,
+  for one reason or another, are entirely dependent on Uberon to produce them.
+  One example of such a project is Ubergraph.
+
+  Another limitation is that the script only works with the OBO format, both
+  for its input and its output.}
+
 \end{frame}
@@
   \end{columns}
+
+  \note{Finally, here is an outline of the entire integration process, what
+  happens when we do a Uberon release. From the main Uberon source file, we
+  create the bridge files using the aforementioned script, and then we merge
+  together: (1) Uberon itself, (2) the bridge files, and (3) the
+  species-specific ontologies, to produce the integrated, multi-species
+  ontology.
+
+  While the process looks superficially simple (though it is, in fact, quite
+  complex, it's just that most of the complexity is hidden behind that
+  script), it is also severely limited.
+
+  It is inflexible, in that the only possible source of mappings to generate
+  the bridge files are the xrefs in Uberon itself -- we cannot integrate
+  mappings coming from another source (which means we cannot delegate the
+  maintenance of the mappings to the species-specific ontologies). Also, we
+  cannot generate a single bridge file on-demand -- the script can only be
+  used to generate all the bridge files in one step. It does not allow to
+  publish the mappings themselves.
+
+  It cannot be applied to other ontologies. There might be other projects that
+  need to integrate several ontologies together, using mappings to connect
+  terms across ontologies, but those projects cannot benefit from the work
+  done in Uberon.}
+
 \end{frame}
@@
   \end{enumerate}
+
+  \note{Here are the key principles that dictated the design of the new
+  approach.
+
+  First, we want a clear separation between the \emph{mappings} and the
+  \emph{bridges} that are derived from the mappings. In our current system,
+  the mappings only exist as annotations within the ontology, and are not
+  really treated as ``first-class objects''. In our new system, we want to
+  focus on the mappings rather than the bridges, because the mappings are the
+  actual informations that we are maintaining -- the bridges are merely a
+  derivatiom from them. In particular, we want to be able to publish the
+  mappings themselves, independently of the bridges.
+
+  Second, we want to store the mappings using a format that is both standard
+  and specifically intended for representing mappings. That format should,
+  among other things, make it possible to store mapping-specific metadata (who
+  asserted the mapping, etc.), make it possible to give precise meaning to a
+  mapping (not just saying that two terms are mapped, but what is the nature
+  of the link between them).
+
+  Lastly, we want a bridge generation process that is much simpler than the
+  current one, and that is not tied to Uberon in any way, so that it can be
+  reused by other projects that have a similar need for integrating several
+  ontologies together.}
+
 \end{frame}
@@
   \end{block}
+
+  \note{Over the past three years, independently of Uberon, a new standard has
+  been devised, specifically for the purpose of representing semantic
+  mappings: the Simple Standard for Sharing Ontological Mappings.
+
+  The SSSOM standard defines a common data model (formalised in LinkML) to
+  represent and manipulate mappings in programming languages. The model
+  notably allows to capture a rich set of metadata about each mapping.
+
+  The standard also defines several serialisation formats to store and
+  exchange mappings, the most important of which being the ``SSSOM/TSV''
+  format, which as the name suggests is based on the ubiquitous TSV format,
+  and is intended to make reusing mappings across resources as easy as
+  possible.
+
+  You see here an very basic example of what a mapping set in SSSOM/TSV format
+  looks like: one mapping per line, each mapping consisting, as its core, of a
+  subject, predicate, object triple. In this example, for simplicitly we only
+  show one bit of metadata (the mapping justification).}
+
 \end{frame}
@@
   \end{center}
+
+  \note{One of the benefits of the SSSOM standard is that it does not simply
+  allow to say that two terms are mapped, or somehow related, to each other.
+  As we've seen in the last example, it also allows to specify an explicit
+  mapping predicate, so as to give a precise meaning to each mapping. It is
+  then possible to distinguish, for example between mappings of terms that
+  represent exactly the same concept in two different vocabularies, and
+  mappings of terms that represent \emph{similar}, but not exactly identical,
+  concepts.
+
+  In fact, as part of the efforts around the SSSOM standard, we have compiled
+  a dedicated vocabulary, the \emph{Semantic Mapping Vocabulary}, to host,
+  among other things, various mapping predicates that can be used to describe
+  mappings. In that vocabulary, as part of this project, we have added a
+  handful of predicates to specifically represent \emph{cross-species
+  mappings}.
+
+  For example, \emph{crossSpeciesExactMatch}, which is intended for mappings
+  between terms that represent the same concept, modulo a taxonomic
+  restriction. We can use that predicate to explicitly say that the FBbt term
+  for the fly heart represents exactly the same concept as the Uberon term for
+  dorsal vessel heart, \emph{except} for the fact that the former is specific
+  to flies.}
+
 \end{frame}
@@
   }
+
+  \note<1>{Now that we can store mappings in a dedicated format, we need a way
+  to use those mappings when we want to integrate several ontologies together.
+  And, as stated in our goals, we want to do that in a way that is, as much as
+  possible, not specific to Uberon, but reusable by other projects as well.
+
+  For that, we have designed, and we propose, a small domain-specific language
+  specifically intended to process SSSOM mappings and turn them into arbitrary
+  OWL axioms. We call that language \emph{SSSOM/T-OWL}, or ``SSSOM Transform
+  to OWL''.
+
+  The core concept of SSSOM/T-OWL is the mapping processing rule, which is
+  made of two elements.
+
+  First, a filter that takes a mapping and returns a boolean value, indicating
+  whether the rule should apply to that mapping or not; the filter can examine
+  all the metadata of a mapping to make its decision.
+
+  Second, either a preprocessor that takes a mapping and returns a modified
+  version of that mapping (allowing to modify a mapping on the fly), or a
+  generator that takes a mapping and returns an OWL axiom.}
+
+  \note<2>{To illustrate, here are some examples of SSSOM/T-OWL rules.
+
+  First, a simple rule with a simple filter that selects all mappings where
+  the subject is a Uberon term, and apply an inversion preprocessor to them.
+  Inversion flips a mapping around the predicate, so that the subject becomes
+  the object and the other way round. So after applying this rule, all
+  mappings with a Uberon subject in the set will become mappings with a Uberon
+  object.
+
+  Second, a rule with an equally simple filter that selects all mappings that
+  use a \emph{skos:broadMatch} predicate, and generate from them a SubClassOf
+  axiom that states that the subject term is a subclass of the object term
+  (which represents a broader concept).
+
+  Third, a more complex rule, with a compound filter that selects mappings
+  that: (1) have a FBbt subject, (2) use a \emph{crossSpeciesExactMatch}
+  predicate, (3) have either been manually curated, or have a confidence level
+  higher than 80\%. When such a mapping is found, the rule creates an
+  equivalence axiom stating that the subject (the FBbt term) is equivalent to
+  the object that is part of the \emph{Drosophila melanogaster} taxon.}
+
 \end{frame}
@@
   \end{block}
+
+  \note{To actually implement our new approach, we first, as a preliminary
+  step, added plugin support to ROBOT, the main tool we use to manipulate OWL
+  and OBO ontologies. This is so that we can quickly add new commands to
+  ROBOT. We then developed a SSSOM plugin for ROBOT, which provides two
+  distinct commands.
+
+  The first is a command to facilitate the transition to SSSOM-maintained
+  mappings by automatically extracting mappings stored as cross-references
+  within an ontology into a mapping set in the SSSOM/TSV format.
+
+  More importantly, the second is a command that implements the SSSOM/T-OWL
+  domain-specific language. That command takes a mapping set and a set of
+  SSSOM/T-OWL rules as input, and will inject into the current ontology the
+  axioms produced by applying the SSSOM/T-OWL rules to the mapping set.}
+
 \end{frame}
@@
   \end{center}
+
+  \note{This leads us to our now, SSSOM-based integration pipeline for Uberon,
+  which now has three distinct inputs: (1) Uberon itself, (2) the mappings
+  between Uberon and the taxon-specific ontologies, and (3) the taxon-specific
+  ontologies.
+
+  The mappings are transformed at build-time into OWL bridge files using a
+  custom SSSOM/T-OWL ruleset, that contains rules similar to the one we have
+  seen earlier: a rule that says that cross-species mappings between FBbt and
+  Uberon are to be transformed into equivalence axioms with a taxonomic
+  restriction to the \emph{Drosophila melanogaster} taxon, a rule that says
+  that cross-species mappings between XAO and Uberon are to be transformed
+  into equivalence axioms with a taxonomic restriction to the \emph{Xenopus}
+  taxon, etc.
+
+  After this step, we have only OWL files, which can all be merged using a
+  standard ROBOT merge operation, producing the multi-species ontology that we
+  want.}
+
 \end{frame}
@@
   \end{block}
+
+  \note{For Uberon itself, the main benefit of the new approach is that we now
+  have a pipeline that is both more flexible --~for example, it allows
+  combining mapping sets from different sources as input~-- and more robust.
+
+  For downstream users of Uberon, the mappings between Uberon and the
+  taxon-specific ontologies are now first-class artefacts. They can have their
+  own metadata, allowing for traceability and accountability of each
+  individual mapping statement. They can be maintained separately from the
+  Uberon ontology itself, for example to delegate them to the taxon-specific
+  ontologies themselves (something we have already started doing for the
+  mappings with FBbt, which are now maintained by FBbt rather than by Uberon).
+  They are available as a distinct release product, which is findable on its
+  own, in a format that is still relatively new but that we hope will become
+  the standard format for semantic mappings.
+
+  And for everyone else, we now have a generic framework to integrate multiple
+  ontologies with SSSOM-maintained mappings. The SSSOM/T-OWL language, and its
+  implementation in the SSSOM ROBOT plugin, has no ties whatsoever to Uberon
+  and can be reused by any project. Only the SSSOM/T-OWL rules that we use are
+  tailored to the needs of Uberon.}
+
 \end{frame}
@@
   \end{block}
+
+  \note{Lastly, here is an example to illustrate some of the benefits of the
+  approach for third-party users. As I've mentioned, at FlyBase we are now
+  maintaining ourselves the mappings between FBbt and Uberon.
+
+  Because FBbt and Uberon are developed separately, there is always a risk
+  that, whenever something changes in either FBbt, Uberon, or the
+  FBbt-to-Uberon mappings, an inter-ontology inconsistency is introduced.
+  Testing to detect and prevent such inconsistencies is difficult and,
+  historically, has never actually been done on a regular basis.
+
+  But with the SSSOM approach, such testing can be done with a single ROBOT
+  command in three steps, as shown here: (1) we merge the current versions of
+  FBbt, Uberon, and CL; (2) we derive OWL axioms from the FBbt-to-Uberon/CL
+  mappings and inject them into the merged ontology; (3) we reason over the
+  merged ontology to detect inconsistencies. This takes but a couple of
+  minutes, and that test is now part of our test suite in FBbt. Thanks to this
+  approach, over the past two years we have rooted out all existing
+  inconsistencies between FBbt and Uberon.}
+
 \end{frame}

### 2024-06-06 presentation.tex
Shorten the section on the previous system.
The presentation should fit in 15 minutes, so we can't afford to spend 3
slides explaining how the previous integration system was working.

@@
 
-  This talk is about how this integration process works.}
+  This talk is about how this integration process works, and how we recently
+  overhauled it to make it easier to maintain, more robust, and FAIRer.}
 
@@
 
-\section{Current system}
+\section{Limits of the previous system}
 
 \begin{frame}
-  \frametitle{Maintaining mappings as cross-references}
-
-  \begin{block}{}\ttfamily
-[Term]
-
-id: UBERON:0015230
-
-name: dorsal vessel heart
-
-def: "The caudal, pulsatile region of the dorsal vessel of the arthropod circulatory system."
-
-is\_a: UBERON:0007100 ! primary circulatory organ
-
-xref: FBbt:00003154
-  \end{block}
-
-  Limitations:
-  \begin{itemize}
-    \item Overloading of the \emph{oboInOwl:hasDbXref} annotation
-    \item No metadata
-    \item Mappings are embedded within the ontology and must be extracted
-  \end{itemize}
-
-  \note{In the current system, mappings are represented and maintained as
-  \emph{cross-references}, that is, as a \emph{oboInOwl:hasDbXref} annotation
-  that is carried by the taxon-neutral term in Uberon and that points to the
-  taxon-specific term in another ontology -- as shown here, where Uberon's
-  ``dorsal vessel heart'' term carries a ``xref'' to FBbt:00003154, which is
-  FBbt's term for the fly heart that we have seen before.
-
-  That's a pretty simple way of representing mappings, but it has several
-  drawbacks.
-
-  First, the \emph{oboInOwl:hasDbXref} annotation has been used for a lot of
-  things since its introduction, and no longer has a clear and unambiguous
-  meaning -- what it means depends on where it is used.
-
-  Second, all this representation allows us to say is that there is a mapping,
-  but we can't know anything else about said mapping. Why does that mapping
-  exist? Who asserted it? When? On what basis? This is simply not recorded.
-
-  Third, the mappings are not available separately from the entire ontology.
-  This makes reusing those mappings more difficult than it should be
-  (difficult to Find, Access and Reuse).}
-
-\end{frame}
-
-\begin{frame}
-  \frametitle{From cross-references to bridge files}
-
-  \begin{columns}
-    \begin{column}{.3\textwidth}
-      \begin{tikzpicture}
-        \node [File] (uberon) at (0,0)     {uberon.obo};
-        \node [File] (script) at (0,-1)    {make-bridge-ontology-from-xrefs.pl};
-        \node [File] (bridge) at (0,-2)    {uberon-bridge-to-fbbt.obo};
-        \node [File]          at (0, -2.5) {uberon-bridge-to-wbbt.obo};
-        \node [File]          at (0, -3)   {uberon-bridge-to-ma.obo};
-        \node [File]          at (0, -3.5) {uberon-bridge-to-....obo}; 
-        \draw[->] (uberon) -- (script) (script) -- (bridge);
-      \end{tikzpicture}
-    \end{column}
-    \begin{column}{.68\textwidth}
-      \begin{block}{uberon-bridge-to-fbbt.obo}\ttfamily
-[Term]
-
-id: FBbt:00003154
-
-intersection\_of: UBERON:0015230
-
-intersection\_of: part\_of NCBITaxon:7227
-      \end{block}
-
-      Can be used to integrate FBbt into Uberon
-    \end{column}
-  \end{columns}
-
-  \medskip
-  Limitations:
-  \begin{itemize}
-    \item Highly Uberon-specific code -- cannot be reused outside of Uberon
-    \item Dependent on the OBO format
-  \end{itemize}
-
-  \note{Whenever we make a new Uberon release, the mappings (stored as
-  cross-references within Uberon) are extracted by a specialised Perl script
-  to produce a set of \emph{bridge files} (one for each species-specific
-  ontology). The bridge files are a OBO materialisation of the mappings; they
-  contain minimalist frames that serve to rattach a species-specific term to a
-  core Uberon term, as in this example, which states that FBbt:00003154 (the
-  fly heart) is equivalent to a Uberon's ``dorsal vessel heart'' that is part
-  of the \emph{Drosophila melanogaster} taxon.
-
-  The main problem with this derivation is that the aforementioned script is
-  highly specific to Uberon. Most of its behaviours are hard-coded, and the
-  script is virtually impossible to re-use outside of Uberon. This means,
-  among other things, that third-party projects that might need those bridges,
-  for one reason or another, are entirely dependent on Uberon to produce them.
-  One example of such a project is Ubergraph.
-
-  Another limitation is that the script only works with the OBO format, both
-  for its input and its output.}
-
-\end{frame}
-
-\begin{frame}
-  \frametitle{Integration pipeline}
+  \frametitle{Overview of the previous integration system}
 
@@
     \end{column}
-    \begin{column}{.55\textwidth}
-      Limitations:
-      \begin{itemize}
-        \item Lack of flexibility
-          \begin{itemize}
-            \item Cannot integrate foreign mappings
-            \item Cannot generate a single bridge file
-          \end{itemize}
-        \item Mappings not available independently
-        \item Not generalizable to other ontologies
-      \end{itemize}
+    \begin{column}{.53\textwidth}
+      \begin{block}{Mappings maintained as xrefs}\scriptsize\ttfamily
+[Term]
+
+id: UBERON:0015230
+
+name: dorsal vessel heart
+
+is\_a: UBERON:0007100 ! primary circulatory organ
+
+xref: FBbt:00003154
+      \end{block}
+
+      \begin{block}{Xrefs transformed to bridge files}\scriptsize\ttfamily
+[Term]
+
+id: FBbt:00003154
+
+intersection\_of: UBERON:0015230
+
+intersection\_of: part\_of NCBITaxon:7227
+      \end{block}
     \end{column}
@@
 
-  \note{Finally, here is an outline of the entire integration process, what
-  happens when we do a Uberon release. From the main Uberon source file, we
-  create the bridge files using the aforementioned script, and then we merge
-  together: (1) Uberon itself, (2) the bridge files, and (3) the
-  species-specific ontologies, to produce the integrated, multi-species
-  ontology.
+  \begin{itemize}
+    \item Mappings not available to third parties
+    \item Pipeline highly Uberon-specific, inflexible and not reusable
+  \end{itemize}
 
-  While the process looks superficially simple (though it is, in fact, quite
-  complex, it's just that most of the complexity is hidden behind that
-  script), it is also severely limited.
+  \note{Mappings are represented and maintained as \emph{cross-references}
+  that are carried by the taxon-neutral terms in Uberon and that point to the
+  taxon-specific terms in other ontologies -- as shown here, where Uberon's
+  ``dorsal vessel heart'' term carries a ``xref'' to FBbt:00003154, which is
+  FBbt's term for the fly heart.
 
-  It is inflexible, in that the only possible source of mappings to generate
-  the bridge files are the xrefs in Uberon itself -- we cannot integrate
-  mappings coming from another source (which means we cannot delegate the
-  maintenance of the mappings to the species-specific ontologies). Also, we
-  cannot generate a single bridge file on-demand -- the script can only be
-  used to generate all the bridge files in one step. It does not allow to
-  publish the mappings themselves.
+  When building Uberon, those cross-references are extracted by a custom
+  script that turns them into \emph{bridge files}, which contains OWL
+  materialisations of the mappings, in the form of equivalence axioms (for
+  example here, an axiom stating the FBbt's ``heart'' is equivalent to an
+  Uberon's ``dorsal vessel heart'' that is part of the \emph{Drosophila
+  melanogaster} taxon).
 
-  It cannot be applied to other ontologies. There might be other projects that
-  need to integrate several ontologies together, using mappings to connect
-  terms across ontologies, but those projects cannot benefit from the work
-  done in Uberon.}
+  This system has severe limitations, one of the most important being that the
+  mappings are directly embedded within the ontology itself and are not
+  independently available. In FAIR terms, this makes the mappings much more
+  difficult to find, access and reuse than we would like.
+
+  Furthermore, the custom script that generates the bridge files from the
+  cross-references is highly specific to Uberon; its behaviour are completely
+  hardcoded to fit Uberon's use case, so the integration logic cannot possibly
+  be reused by other projects.}
 
@@
 
-\section{New approach}
-
 \begin{frame}
-  \frametitle{Principles}
+  \frametitle{Goals for a new approach}
 
   \begin{enumerate}
-    \item Clear separation between
+    \item Cross-ontology mappings (COM) as first-class entities
       \begin{itemize}
-        \item cross-ontology \emph{mappings} (COM)
-        \item cross-ontology \emph{bridges} (COB)
+        \item available independently of the ontology itself
       \end{itemize}
-    \item Represent COM using a standard format intended for that purpose
+    \item Represent mappings using a standard format intended for that purpose
       \begin{itemize}
@@
       \end{itemize}
-    \item Make it easy to derive COB from COM
+    \item Make it easy to derive OWL bridge files from mappings
       \begin{itemize}
@@
 
-  \note{Here are the key principles that dictated the design of the new
-  approach.
+  \note{In response to the limits of the previous system, here are the key
+  principles that dictated the design of the new approach.
 
-  First, we want a clear separation between the \emph{mappings} and the
-  \emph{bridges} that are derived from the mappings. In our current system,
-  the mappings only exist as annotations within the ontology, and are not
-  really treated as ``first-class objects''. In our new system, we want to
-  focus on the mappings rather than the bridges, because the mappings are the
-  actual informations that we are maintaining -- the bridges are merely a
-  derivatiom from them. In particular, we want to be able to publish the
-  mappings themselves, independently of the bridges.
+  First, we want to treat mappings as ``first-class entities'', on the same
+  level as the terms of the ontology. Among other things, this means that they
+  must be publicly available on their own, without requiring people interested
+  in them to forcefully extract them from the ontology.
 
@@
   and specifically intended for representing mappings. That format should,
-  among other things, make it possible to store mapping-specific metadata (who
-  asserted the mapping, etc.), make it possible to give precise meaning to a
-  mapping (not just saying that two terms are mapped, but what is the nature
-  of the link between them).
+  among other things, make it possible to store mapping-specific metadata, and
+  to give precise meaning to a mapping (not just saying that two terms are
+  mapped, but what is the nature of the link between them).
 
@@
 \end{frame}
+
+\section{Implementation}
 
@@
 
-  \note{This leads us to our now, SSSOM-based integration pipeline for Uberon,
+  \note{This leads us to our new, SSSOM-based integration pipeline for Uberon,
   which now has three distinct inputs: (1) Uberon itself, (2) the mappings

### 2024-06-14 presentation.tex
Move some slides to the appendix.

@@
   multi-species ontology that covers all our model organisms. For the past 12
-  years or so, this has been provided by Uberon, the Uber Anatomy Ontology.
+  years or so, this has been the goal of Uberon, the Uber Anatomy Ontology.
 
-  Rather than duplicating the work already done by pre-existing,
-  species-specific ontologies, Uberon's strategy is to provide a taxon-neutral
-  core framework, describing common anatomical structures in broadly generic
-  terms, and to plug the species-specific ontologies into that core.}
+  Uberon's strategy to achieve that goal is to provide a taxon-neutral core
+  framework, describing common anatomical structures in broadly generic terms,
+  and to plug the species-specific ontologies into that core.}
 
@@
 
-  This is the role of cross-ontology \emph{mappings} -- basically, a mapping
-  is a statement that one term from an ontology A is ``mapped'', or
-  ``related'', to another term in an ontology B.}
+  This is the role of cross-ontology \emph{mappings}.}
 
@@
   cross-references is highly specific to Uberon; its behaviour are completely
-  hardcoded to fit Uberon's use case, so the integration logic cannot possibly
-  be reused by other projects.}
+  hardcoded to fit Uberon's use case, so the integration logic is difficult to
+  modify and almost impossible to reuse by other projects.}
 
@@
   First, we want to treat mappings as ``first-class entities'', on the same
-  level as the terms of the ontology. Among other things, this means that they
-  must be publicly available on their own, without requiring people interested
-  in them to forcefully extract them from the ontology.
+  level as the terms of the ontology. This notably means that they must be
+  publicly available on their own, without requiring people interested in them
+  to forcefully extract them from the ontology.
 
   Second, we want to store the mappings using a format that is both standard
-  and specifically intended for representing mappings. That format should,
-  among other things, make it possible to store mapping-specific metadata, and
-  to give precise meaning to a mapping (not just saying that two terms are
-  mapped, but what is the nature of the link between them).
+  and specifically intended for representing mappings. That format should make
+  it possible to store mapping-specific metadata, and to give precise meaning
+  to a mapping (not just saying that two terms are mapped, but: who asserted
+  that those two terms should be mapped? When? On what basis? And what does
+  the mapping actually means?).
 
@@
   \note{One of the benefits of the SSSOM standard is that it does not simply
-  allow to say that two terms are mapped, or somehow related, to each other.
-  As we've seen in the last example, it also allows to specify an explicit
-  mapping predicate, so as to give a precise meaning to each mapping. It is
-  then possible to distinguish, for example between mappings of terms that
-  represent exactly the same concept in two different vocabularies, and
-  mappings of terms that represent \emph{similar}, but not exactly identical,
-  concepts.
+  allow to say that two terms are mapped, to each other.  As we've seen in the
+  last example, it also allows to specify an explicit mapping predicate, so as
+  to give a precise meaning to each mapping. It is then possible to
+  distinguish, for example between mappings of terms that represent exactly
+  the same concept in two different vocabularies, and mappings of terms that
+  represent \emph{similar}, but not exactly identical, concepts.
 
-  In fact, as part of the efforts around the SSSOM standard, we have compiled
-  a dedicated vocabulary, the \emph{Semantic Mapping Vocabulary}, to host,
-  among other things, various mapping predicates that can be used to describe
-  mappings. In that vocabulary, as part of this project, we have added a
-  handful of predicates to specifically represent \emph{cross-species
-  mappings}.
+  The authors of the SSSOM standard have compiled a dedicated vocabulary, the
+  \emph{Semantic Mapping Vocabulary}, to host, among other things, various
+  mapping predicates that can be used to describe mappings. In that
+  vocabulary, as part of this project, we have added a handful of predicates
+  to specifically represent \emph{cross-species mappings}.
 
-  For example, \emph{crossSpeciesExactMatch}, which is intended for mappings
-  between terms that represent the same concept, modulo a taxonomic
-  restriction. We can use that predicate to explicitly say that the FBbt term
-  for the fly heart represents exactly the same concept as the Uberon term for
-  dorsal vessel heart, \emph{except} for the fact that the former is specific
-  to flies.}
+  The most important being \emph{crossSpeciesExactMatch}, which is intended
+  for mappings between terms that represent the same concept, modulo a
+  taxonomic restriction. We can use that predicate to explicitly say that the
+  FBbt term for the fly heart represents exactly the same concept as the
+  Uberon term for dorsal vessel heart, \emph{except} for the fact that the
+  former is specific to flies.}
 
@@
   to use those mappings when we want to integrate several ontologies together.
-  And, as stated in our goals, we want to do that in a way that is, as much as
+  And, as per our third goal, we want to do that in a way that is, as much as
   possible, not specific to Uberon, but reusable by other projects as well.
@@
 \begin{frame}
+  \frametitle{New SSSOM-based integration pipeline}
+
+  \begin{center}
+    \begin{tikzpicture}[node distance=.5cm]
+      \node [Entity] (uberon) {Uberon};
+      \node [Entity,right=of uberon] (com) {Mapping sets};
+      \node [Entity,right=of com,align=left] (ssao) {Taxon-specific\\
+                                                     ontologies};
+      \node [Process,below=of com] (sssomt) {SSSOM/T-OWL};
+      \node [Entity,below=of sssomt] (cob) {Bridge files};
+      \node [Process,below=of cob] (merge) {OWL merge};
+      \node [Entity,below=of merge] (cm) {Multi-species ontology};
+      \draw [->] (com) -- (sssomt) (sssomt) -- (cob);
+      \draw [->] (cob) -- (merge) (merge) -- (cm);
+      \draw [->] (uberon.south) |- (merge.west);
+      \draw [->] (ssao.south) |- (merge.east);
+    \end{tikzpicture}
+  \end{center}
+
+  \note{This leads us to our new, SSSOM-based integration pipeline for Uberon.
+  That pipeline has three distinct inputs: (1) Uberon itself, (2) the mappings
+  between Uberon and the taxon-specific ontologies (which are no longer
+  embedded within the ontology itself), and (3) the taxon-specific ontologies.
+
+  The mappings are transformed at build-time into OWL bridge files using a
+  custom SSSOM/T-OWL ruleset, that contains rules similar to the one we have
+  seen earlier: a rule that says that cross-species mappings between FBbt and
+  Uberon are to be transformed into equivalence axioms with a taxonomic
+  restriction to the \emph{Drosophila melanogaster} taxon, a rule that says
+  that cross-species mappings between XAO and Uberon are to be transformed
+  into equivalence axioms with a taxonomic restriction to the \emph{Xenopus}
+  taxon, etc.
+
+  That step is performed by a dedicated ROBOT plugins that implement the
+  SSSOM/T-OWL language.
+
+  After this step, we have only OWL files, which can all be merged using a
+  standard ROBOT merge operation, producing the multi-species ontology that we
+  want.}
+
+\end{frame}
+
+\section{Conclusion}
+
+\begin{frame}
+  \frametitle{Benefits of the new approach}
+
+  \begin{block}{Cross-ontology mappings as first-class artefacts}
+    \begin{itemize}
+      \item With their own metadata
+      \item Can be maintained separately
+      \item Available as a separate release product in a standard format
+    \end{itemize}
+  \end{block}
+
+  \begin{block}{Generic framework for SSSOM-based integration of multiple ontologies}
+    \begin{itemize}
+      \item SSSOM/T-OWL is completely independent of Uberon
+      \item Only the ruleset is specific
+    \end{itemize}
+  \end{block}
+
+  \note{For Uberon itself, the main benefit of the new approach is that we now
+  have a pipeline that is both more flexible --~for example, easy to modify,
+  as all the logic is in the SSSOM/T-OWL ruleset~-- and more robust.
+
+  For downstream users of Uberon, the mappings between Uberon and the
+  taxon-specific ontologies are now first-class artefacts. They can have their
+  own metadata, allowing for traceability and accountability of each
+  individual mapping statement. They can be maintained separately from the
+  Uberon ontology itself, for example to delegate them to the taxon-specific
+  ontologies themselves (something we have already started doing for the
+  mappings with FBbt, which are now maintained by FBbt rather than by Uberon).
+  They are available as a distinct release product, which is findable on its
+  own, in a format that is still relatively new but that we hope will become
+  the standard format for semantic mappings.
+
+  And for everyone else, we now have a generic framework to integrate multiple
+  ontologies with SSSOM-maintained mappings. The SSSOM/T-OWL language, and its
+  implementation in the SSSOM ROBOT plugin, has no ties whatsoever to Uberon
+  and can be reused by any project. Only the SSSOM/T-OWL rules that we use are
+  tailored to the needs of Uberon.}
+
+\end{frame}
+
+\appendix
+\section{\appendixname}
+
+\begin{frame}
   \frametitle{Implementation in ROBOT commands}
@@
   axioms produced by applying the SSSOM/T-OWL rules to the mapping set.}
-
-\end{frame}
-
-\begin{frame}
-  \frametitle{New SSSOM-based integration pipeline}
-
-  \begin{center}
-    \begin{tikzpicture}[node distance=.5cm]
-      \node [Entity] (uberon) {Uberon};
-      \node [Entity,right=of uberon] (com) {Mapping sets};
-      \node [Entity,right=of com,align=left] (ssao) {Taxon-specific\\
-                                                     ontologies};
-      \node [Process,below=of com] (sssomt) {SSSOM/T-OWL};
-      \node [Entity,below=of sssomt] (cob) {Bridge files};
-      \node [Process,below=of cob] (merge) {OWL merge};
-      \node [Entity,below=of merge] (cm) {Multi-species ontology};
-      \draw [->] (com) -- (sssomt) (sssomt) -- (cob);
-      \draw [->] (cob) -- (merge) (merge) -- (cm);
-      \draw [->] (uberon.south) |- (merge.west);
-      \draw [->] (ssao.south) |- (merge.east);
-    \end{tikzpicture}
-  \end{center}
-
-  \note{This leads us to our new, SSSOM-based integration pipeline for Uberon,
-  which now has three distinct inputs: (1) Uberon itself, (2) the mappings
-  between Uberon and the taxon-specific ontologies, and (3) the taxon-specific
-  ontologies.
-
-  The mappings are transformed at build-time into OWL bridge files using a
-  custom SSSOM/T-OWL ruleset, that contains rules similar to the one we have
-  seen earlier: a rule that says that cross-species mappings between FBbt and
-  Uberon are to be transformed into equivalence axioms with a taxonomic
-  restriction to the \emph{Drosophila melanogaster} taxon, a rule that says
-  that cross-species mappings between XAO and Uberon are to be transformed
-  into equivalence axioms with a taxonomic restriction to the \emph{Xenopus}
-  taxon, etc.
-
-  After this step, we have only OWL files, which can all be merged using a
-  standard ROBOT merge operation, producing the multi-species ontology that we
-  want.}
-
-\end{frame}
-
-\section{Conclusion}
-
-\begin{frame}
-  \frametitle{Benefits of the new approach}
-
-  \begin{block}{Cross-ontology mappings as first-class artefacts}
-    \begin{itemize}
-      \item With their own metadata
-      \item Can be maintained separately
-      \item Available as a separate release product in a standard format
-    \end{itemize}
-  \end{block}
-
-  \begin{block}{Generic framework for SSSOM-based integration of multiple ontologies}
-    \begin{itemize}
-      \item SSSOM/T-OWL is completely independent of Uberon
-      \item Only the ruleset is specific
-    \end{itemize}
-  \end{block}
-
-  \note{For Uberon itself, the main benefit of the new approach is that we now
-  have a pipeline that is both more flexible --~for example, it allows
-  combining mapping sets from different sources as input~-- and more robust.
-
-  For downstream users of Uberon, the mappings between Uberon and the
-  taxon-specific ontologies are now first-class artefacts. They can have their
-  own metadata, allowing for traceability and accountability of each
-  individual mapping statement. They can be maintained separately from the
-  Uberon ontology itself, for example to delegate them to the taxon-specific
-  ontologies themselves (something we have already started doing for the
-  mappings with FBbt, which are now maintained by FBbt rather than by Uberon).
-  They are available as a distinct release product, which is findable on its
-  own, in a format that is still relatively new but that we hope will become
-  the standard format for semantic mappings.
-
-  And for everyone else, we now have a generic framework to integrate multiple
-  ontologies with SSSOM-maintained mappings. The SSSOM/T-OWL language, and its
-  implementation in the SSSOM ROBOT plugin, has no ties whatsoever to Uberon
-  and can be reused by any project. Only the SSSOM/T-OWL rules that we use are
-  tailored to the needs of Uberon.}
 

### 2024-06-16 presentation.tex
Add highlighting.
Improve standalone presentation.
Make the presentation easier to follow when the presenter is not around.

Also add an "about" slide.

@@
             circulatory organ <- {
-              primary circulatory organ <- { heart, dorsal vessel heart },
+              primary circulatory organ <- { heart, dorsal vessel heart[fill=GenericHighlight] },
               accessory circulatory organ <- lymph heart
@@
             anatomical structure <-
-            heart <-["P"] {
+            heart[fill=GenericHighlight] <-["P"] {
               aortic funnel, heart chamber, dorsal diaphragm
@@
                 heart,
-                dorsal vessel heart <- "heart (drosophila)" <-["P"] {
+                dorsal vessel heart[fill=GenericHighlight]
+                    <- "heart (drosophila)"[fill=GenericHighlight] <-["P"] {
                   aortic funnel, heart chamber, dorsal diaphragm
@@
 
-xref: FBbt:00003154
+\textcolor{red}{xref: FBbt:00003154}
       \end{block}
@@
       \hline
-      subject\_id   & subject\_label                & predicate\_id                 & object\_id      & mapping\_justification\\
+      subject\_id   & predicate\_id                 & object\_id      & mapping\_justification        & mapping\_date\\
       \hline
-      FBbt:00000001 & organism                      & semapv:crossSpeciesExactMatch & UBERON:0000468  & semapv:ManualMappingCuration\\
-      FBbt:00000002 & tagma                         & semapv:crossSpeciesExactMatch & UBERON:6000002  & semapv:ManualMappingCuration\\
+      FBbt:00000001 & semapv:crossSpeciesExactMatch & UBERON:0000468  & semapv:ManualMappingCuration  & 2024-06-01\\
+      FBbt:00000002 & semapv:crossSpeciesExactMatch & UBERON:6000002  & semapv:ManualMappingCuration  & 2024-06-01\\
       ...           & ...                           & ...             & ...                           & ...\\
-      FBbt:00003154 & heart                         & semapv:crossSpeciesExactMatch & UBERON:0015230  & semapv:ManualMappingCuration\\
-      ...           & ...                           &...                           & ...             & ...\\
+      \textcolor<2>{red}{FBbt:00003154} & \textcolor<2>{red}{semapv:crossSpeciesExactMatch} & \textcolor<2>{red}{UBERON:0015230}  & \textcolor<3>{red}{semapv:ManualMappingCuration}  & \textcolor<3>{red}{2024-06-01}\\
+      ...           & ...                           & ...             & ...                           & ...\\
       \hline
@@
 
-  \note{Over the past three years, independently of Uberon, a new standard has
+  \begin{overlayarea}{\textwidth}{1cm}
+    \only<2>{Mapping proper: subject -- predicate -- object triple}
+    \only<3>{Mapping metadata}
+  \end{overlayarea}
+
+  \note<1>{Over the past three years, independently of Uberon, a new standard has
   been devised, specifically for the purpose of representing semantic
@@
   You see here an very basic example of what a mapping set in SSSOM/TSV format
-  looks like: one mapping per line, each mapping consisting, as its core, of a
-  subject, predicate, object triple. In this example, for simplicitly we only
-  show one bit of metadata (the mapping justification).}
+  looks like: one mapping per line, each mapping consisting, at its core, of a
+  subject, predicate, object triple, associated with several metadata.}
 
@@
 
-  \only<1>{
   \begin{block}{Model of a mapping processing rule}
@@
   \end{block}
-  }
-  \only<2>{
-    \begin{block}{Inverting a mapping}\ttfamily\scriptsize
-subject==UBERON:* -> invert();
-    \end{block}
 
-    \begin{block}{Generating a simple axiom}\ttfamily\scriptsize
-predicate==skos:broadMatch -> create\_axiom("\%subject\_id SubClassOf: \%object\_id");
-    \end{block}
-
-    \begin{block}{Generating a more complex axiom}\ttfamily\scriptsize
-subject==FBbt:*\\
-\hspace{1cm} \&\& predicate==semapv:crossSpeciesExactMatch\\
-\hspace{1cm} \&\& (mapping\_justification==semapv:ManualMappingCuration || confidence>=0.8)\\
-->  create\_axiom("\%subject\_id EquivalentTo: \%object\_id and\\
-\hspace{2cm} (BFO:0000050 some NCBITaxon:7227"));
-    \end{block}
-  }
-
-  \note<1>{Now that we can store mappings in a dedicated format, we need a way
+  \note{Now that we can store mappings in a dedicated format, we need a way
   to use those mappings when we want to integrate several ontologies together.
@@
   generator that takes a mapping and returns an OWL axiom.}
+\end{frame}
 
-  \note<2>{To illustrate, here are some examples of SSSOM/T-OWL rules.
+\begin{frame}
+  \frametitle{From SSSOM sets to bridge files: SSSOM/T-OWL}
+
+  Examples of SSSOM/T-OWL rules
+
+  \begin{overlayarea}{\textwidth}{3cm}
+    \only<1-2>{
+      \begin{block}{Inverting a mapping}\ttfamily\scriptsize
+        \textcolor<1>{red}{subject==UBERON:*} -> \textcolor<2>{red}{invert()};
+      \end{block}
+    }
+    \only<3-4>{
+      \begin{block}{Generating a simple axiom}\ttfamily\scriptsize
+        \textcolor<3>{red}{predicate==skos:broadMatch} -> \textcolor<4>{red}{create\_axiom("\%subject\_id SubClassOf: \%object\_id")};
+      \end{block}
+    }
+    \only<5-10>{
+      \begin{block}{Generating a more complex axiom}\ttfamily\scriptsize
+        \textcolor<5>{red}{subject==FBbt:*}\\
+        \hspace{1cm} \&\& \textcolor<6>{red}{predicate==semapv:crossSpeciesExactMatch}\\
+        \hspace{1cm} \&\& (\textcolor<7>{red}{mapping\_justification==semapv:ManualMappingCuration} \textcolor<8>{red}{||} \textcolor<9>{red}{confidence>=0.8)}\\
+        ->  \textcolor<10>{red}{create\_axiom("\%subject\_id EquivalentTo: \%object\_id and\\
+        \hspace{2cm} (BFO:0000050 some NCBITaxon:7227"))};
+      \end{block}
+    }
+  \end{overlayarea}
+
+  \begin{overlayarea}{\textwidth}{2.5cm}
+    \only<1-2>{\textcolor<1>{red}{Select mappings with a UBERON subject}
+
+    and \textcolor<2>{red}{flip them around the predicate}}
+    \only<3-4>{\textcolor<3>{red}{Select mappings with a \emph{skos:broadMatch} predicate}
+
+    and \textcolor<4>{red}{create a \emph{SubClassOf} axiom}}
+    \only<5-10>{Select mappings
+      \begin{itemize}
+        \item \textcolor<5>{red}{with a FBbt subject}
+        \item \textcolor<6>{red}{with a \emph{semapv:crossSpeciesExactMatch} predicate}
+        \item that have been \textcolor<7>{red}{manually curated} \textcolor<8>{red}{or} have a \textcolor<9>{red}{confidence higher than 80\%}
+      \end{itemize}
+
+      and \textcolor<10>{red}{create an equivalence axiom with a taxonomic restriction}}
+  \end{overlayarea}
+
+  \note<1>{To illustrate, here are some examples of SSSOM/T-OWL rules.
 
@@
 
+\begin{frame}
+  \frametitle{About}
+
+  \begin{block}{Acknowledgements}
+    \begin{itemize}
+      \item Grant BB/T014008 from BBSRC (UK) and NSF/BIO (US)
+      \item Mapping Commons project for the SSSOM standard
+    \end{itemize}
+  \end{block}
+
+  \begin{block}{\includegraphics[scale=.26]{ccbysa}\hspace{.2em}2024 Damien Goutte-Gattat}
+    This work is licensed under the Creative Commons Attributions-ShareAlike
+    4.0 License.
+  \end{block}
+
+  \begin{block}{Contact}
+    \href{mailto:dpg44@cam.ac.uk}{\texttt{dpg44@cam.ac.uk}}
+  \end{block}
+\end{frame}
+
 \appendix
-\section{\appendixname}
+%\section{\appendixname}
 
-\begin{frame}
+\begin{frame}[noframenumbering]
   \frametitle{Implementation in ROBOT commands}
@@
 
-\begin{frame}[fragile]
+\begin{frame}[fragile,noframenumbering]
   \frametitle{Example: FBbt/Uberon/CL integration test}

### 2024-06-16 presentation.tex
Add the revision number to the "about" slide.

@@
         \hspace{1cm} \&\& \textcolor<6>{red}{predicate==semapv:crossSpeciesExactMatch}\\
-        \hspace{1cm} \&\& (\textcolor<7>{red}{mapping\_justification==semapv:ManualMappingCuration} \textcolor<8>{red}{||} \textcolor<9>{red}{confidence>=0.8)}\\
+        \hspace{1cm} \&\&
+        (\textcolor<7>{red}{mapping\_justification==semapv:ManualMappingCuration}
+        \textcolor<8>{red}{||} \textcolor<9>{red}{confidence>=0.8})\\
         ->  \textcolor<10>{red}{create\_axiom("\%subject\_id EquivalentTo: \%object\_id and\\
@@
 
-  \begin{block}{\includegraphics[scale=.26]{ccbysa}\hspace{.2em}2024 Damien Goutte-Gattat}
-    This work is licensed under the Creative Commons Attributions-ShareAlike
-    4.0 License.
+  \begin{block}{\includegraphics[scale=.26]{ccbysa}\hspace{.2em}%
+    2024 Damien Goutte-Gattat \texttt{<dpg44@cam.ac.uk>}}
+    This presentation is released under the Creative Commons
+    Attributions-ShareAlike 4.0 License.
   \end{block}
 
-  \begin{block}{Contact}
-    \href{mailto:dpg44@cam.ac.uk}{\texttt{dpg44@cam.ac.uk}}
+  \begin{block}{Revision}\ttfamily\scriptsize
+    \input{revision.tex}
   \end{block}

### 2024-06-16 paper.tex
Rename the paper source file.
This is to prepare the merge with the presentation repository.

--- 2024-05-23 sssomt.tex
@@
 either two terms are mapped or they are not, without room to express
-more subtle relations. (iii) Cross-references have no clear and
-universally agreed upon meaning. When two terms are mapped to each
-other, the meaning of the mapping is left unspecified, and typically has
-to be inferred from the ontologies that are being mapped (for example, a
-mapping between the species-neutral Uberon and the
-\species{Drosophila}-specific FBbt can be inferred to be a cross-species
-mapping). This is made worse by the fact that cross-references in OBO
-ontologies are used for many different things beyond just mappings; for
-example, when used to annotate a term definition, cross-references
-typically attribute a source for the definition, akin to a citation in
-an academic paper.
+more subtle relations. (iii) Cross-references are embedded within the
+ontology itself, and are therefore not easily reusable by third parties.
+(iv) Cross-references have no clear and universally agreed upon meaning.
+When two terms are mapped to each other, the meaning of the mapping is
+left unspecified, and typically has to be inferred from the ontologies
+that are being mapped (for example, a mapping between the
+species-neutral Uberon and the \species{Drosophila}-specific FBbt can be
+inferred to be a cross-species mapping). This is made worse by the fact
+that cross-references in OBO ontologies are used for many different
+things beyond just mappings; for example, when used to annotate a term
+definition, cross-references typically attribute a source for the
+definition, akin to a citation in an academic paper.
 
@@
 
-To construct a consistent and maintainable cross-species anatomy
-ontology, we have updated the integration mechanism between Uberon, CL,
-and the species-centric ontologies by using SSSOM to manage the mappings
-betweeen the ontologies. While full transition to SSSOM is not yet
-complete, we now have the needed infrastructure in place to complete
-this migration, and we have a more robust and more flexible pipeline for
-creating the bridges needed for \emph{composite-metazoan}. The logic for
-deriving the bridges is expressed using a simple DSL (SSSOM/T-OWL) that
-most ontology editors should be able to master quickly, should the need
-arise to modify said logic (for example to add a new bridge). While the
-precise SSSOM/T-OWL rules we use are obviously specific to the needs of
-Uberon, we believe the general approach of maintaining mappings as SSSOM
-sets and transforming them into OWL bridges using a simple DSL could be
-used by any project that requires integrating several ontologies
+In this project, we have (i) expanded the SEMAPV vocabulary so that
+cross-species mappings can be specifically represented in SSSOM; (ii)
+introduced a domain-specific language (SSSOM/T-OWL) to allow converting
+SSSOM mappings into arbitrary OWL axioms; (iii) implemented said
+language in a new ROBOT plugin to inject SSSOM-derived axioms into a OWL
+ontology; (iv) used those new tools to update the integration mechanism
+between Uberon, CL, and the species-centric ontologies, with an approach
+centered on the use of SSSOM, rather than cross-references, to manage
+the mappings between the ontologies.
+
+As a result, cross-species mappings across anatomy ontologies are now
+available as a \emph{bona fide} release artefact of Uberon, provided as
+a distinct file in a known location in a standard format, thereby
+greatly facilitating the reuse of said mappings by any interested third
+party. The logic for deriving the bridge files (needed to correctly
+merge Uberon and the species-specific ontologies), now expressed in the
+SSSOM/T-OWL language, is itself consequently more easily maintainable
+and reusable.
+
+While several aspects of this project (most notably the precise
+SSSOM/T-OWL rules we use) are specific to the needs of Uberon, we
+believe the general approach of maintaining mappings as SSSOM sets and
+transforming them into OWL bridges using a simple DSL could be
+generalised to any project that requires integrating several ontologies
 together.

### 2024-06-16 presentation.tex
Update license and add README.
The paper is published under CC-BY, so we'll use CC-BY for the
presentation as well. We add the copyright notices and the full text of
the CC-BY license.

Also add a minimal README.

@@
+% SSSOM/T and Uberon presentation
+% © 2024 Damien Goutte-Gattat
+% 
+% This work is released under the Creative Commons Attribution 4.0
+% International License. To view a copy of this license, visit
+% https://creativecommons.org/licenses/by/4.0/.
 \usepackage{fontspec}
@@
 
-  \begin{block}{\includegraphics[scale=.26]{ccbysa}\hspace{.2em}%
+  \begin{block}{\includegraphics[scale=.26]{ccby}\hspace{.2em}%
     2024 Damien Goutte-Gattat \texttt{<dpg44@cam.ac.uk>}}
-    This presentation is released under the Creative Commons
-    Attributions-ShareAlike 4.0 License.
+    This presentation is released under the Creative Commons Attribution
+    4.0 International License (CC BY 4.0).
   \end{block}

### 2024-06-17 presentation.tex
Make the presentation more self-sufficient.
Add more text and animation to make the presentation reasonably
understandable in the absence of the presenter.

@@
 % TikZ libraries and styles
-\usetikzlibrary{positioning,shapes.geometric,graphs,graphdrawing,quotes,fit}
+\usetikzlibrary{positioning,shapes.geometric,graphs,graphdrawing,quotes,fit,overlay-beamer-styles}
 \usegdlibrary{layered,trees}
@@
       \begin{tikzpicture}[nodes={shape=circle,draw,fill=GenericItem}]
-        \node (core) {Uberon};
-        \node (fbbt) at (150:2cm) {FBbt};
-        \node (zfa)  at (90:2cm)  {ZFA};
-        \node (xao)  at (30:2cm)  {XAO};
-        \node (ma)   at (210:2cm) {MA};
-        \node (wbbt) at (270:2cm) {WBbt};
+        \node (core) [alt=<6>{fill=GenericHighlight}{}] {Uberon};
+        \node (fbbt) [alt=<{1,7}>{fill=GenericHighlight}{}] at (150:2cm) {FBbt};
+        \node (zfa)  [alt=<{2,7}>{fill=GenericHighlight}{}] at (90:2cm)  {ZFA};
+        \node (xao)  [alt=<{3,7}>{fill=GenericHighlight}{}] at (30:2cm)  {XAO};
+        \node (ma)   [alt=<{4,7}>{fill=GenericHighlight}{}] at (210:2cm) {MA};
+        \node (wbbt) [alt=<{5,7}>{fill=GenericHighlight}{}] at (270:2cm) {WBbt};
         \node (tx1) at  (310:2cm) {...};
@@
     \begin{column}{.45\textwidth}
-      Strategy:
+      Many species-specific anatomy ontologies available:
+      \smallskip
+      \begin{overlayarea}{\textwidth}{1cm}\color{red}
+        \only<1>{\emph{Drosophila} Anatomy Ontology}
+        \only<2>{Zebrafish Anatomy Ontology}
+        \only<3>{Xenopus Anatomy Ontology}
+        \only<4>{Mouse Anatomy Ontology}
+        \only<5>{\emph{C. elegans} Anatomy Ontology}
+      \end{overlayarea}
+
+      Strategy for a multi-species ontology:
       \begin{itemize}
-        \item Provide a taxon-neutral core framework
-        \item Integrate existing taxon-specific ontologies
+        \item Provide a \textcolor<6>{red}{taxon-neutral core framework}
+        \item Integrate \textcolor<7>{red}{existing taxon-specific ontologies}
       \end{itemize}
@@
 
-  \note{Anatomy ontologies are used in many aspects of biocuration, for
+  \note<1>{Anatomy ontologies are used in many aspects of biocuration, for
   example to capture the organ of origin of a biological sample, the tissue
@@
 \begin{frame}
-  \frametitle{Example: integration with FBbt}
+  \frametitle{Example: integrating FBbt into Uberon}
 
@@
         \end{tikzpicture}
+
+        FBbt's ``heart'' is the fly-specific equivalent of Uberon's ``dorsal vessel heart''
       \end{column}
@@
       \begin{column}{.3\textwidth}
-        Integrated hierarchy
+        Integrated hierarchy:
+
+        \smallskip
+        FBbt's ``heart'' classified under Uberon's ``dorsal vessel heart''
       \end{column}
@@
   \begin{columns}
-    \begin{column}{.44\textwidth}
-      \begin{tikzpicture}[node distance=.5cm]
-        \node [Entity] (uberon) {Uberon};
-        \node [Process,below right=of uberon] (script) {Custom script};
+    \begin{column}{.46\textwidth}
+      \begin{tikzpicture}[node distance=.3cm]
+        \node [Entity,text width=1.3cm,alt=<1>{fill=GenericHighlight}{}] (uberon) {Uberon\\(incl. xrefs)};
+        \node [Process,below right=of uberon,alt=<3>{fill=GenericHighlight}{}] (script) {Custom script};
         \node [Entity,above right=of script,align=left] (ssao) {Taxon-specific\\
                                                                 ontologies};
-        \node [Entity,below=of script] (cob) {Bridge files};
+        \node [Entity,below=of script,alt=<2>{fill=GenericHighlight}{}] (cob) {Bridge files};
         \node [Process,below=of cob] (merge) {OWL merge};
@@
     \end{column}
-    \begin{column}{.53\textwidth}
-      \begin{block}{Mappings maintained as xrefs}\scriptsize\ttfamily
+    \begin{column}{.51\textwidth}
+      \begin{block}{Mappings maintained as \textcolor<1>{red}{xrefs}}\scriptsize\ttfamily
 [Term]
@@
 
-\textcolor{red}{xref: FBbt:00003154}
+\textcolor<1>{red}{xref: FBbt:00003154}
       \end{block}
 
-      \begin{block}{Xrefs transformed to bridge files}\scriptsize\ttfamily
+      \begin{block}{Xrefs transformed to \textcolor<2>{red}{bridge files}}\scriptsize\ttfamily
 [Term]
@@
     \item Mappings not available to third parties
-    \item Pipeline highly Uberon-specific, inflexible and not reusable
+    \item \textcolor<3>{red}{Custom pipeline} highly Uberon-specific, inflexible and not reusable
   \end{itemize}
 
-  \note{Mappings are represented and maintained as \emph{cross-references}
+  \note<1>{Mappings are represented and maintained as \emph{cross-references}
   that are carried by the taxon-neutral terms in Uberon and that point to the
@@
   \end{center}
+
+  \emph{crossSpeciesExactMatch}: ``a match where the subject is considered analogous
+  to the object in a different taxonomic grouping''
 
@@
 
+  \begin{columns}
+    \begin{column}{.45\textwidth}
       \begin{center}
@@
                                                          ontologies};
-      \node [Process,below=of com] (sssomt) {SSSOM/T-OWL};
+          \node [Process,below=of com,alt=<2>{fill=GenericHighlight}{}] (sssomt) {SSSOM/T-OWL};
           \node [Entity,below=of sssomt] (cob) {Bridge files};
-      \node [Process,below=of cob] (merge) {OWL merge};
+          \node [Process,below=of cob,alt=<3>{fill=GenericHighlight}{}] (merge) {OWL merge};
           \node [Entity,below=of merge] (cm) {Multi-species ontology};
@@
       \end{center}
+    \end{column}
+    \begin{column}{.45\textwidth}
+      \begin{enumerate}
+        \item Generating the bridge files through a \textcolor<2>{red}{SSSOM/T-OWL ruleset}\\
+              (\texttt{robot sssom:inject})
+        \item \textcolor<3>{red}{Merging} Uberon, the bridges, and the taxon-specific ontologies\\
+              (\texttt{robot merge})
+      \end{enumerate}
+    \end{column}
+  \end{columns}
 
-  \note{This leads us to our new, SSSOM-based integration pipeline for Uberon.
+  \note<1>{This leads us to our new, SSSOM-based integration pipeline for Uberon.
   That pipeline has three distinct inputs: (1) Uberon itself, (2) the mappings

### 2024-06-17 presentation.tex
Add intercalary slide before the extra slides.
Fix the handout version.
When generating the handout version of the presentation, use the
`handout` Beamer option and `handout:X` overlay specifications to avoid
generating one slide for every single overlay.

@@
       \begin{tikzpicture}[nodes={shape=circle,draw,fill=GenericItem}]
-        \node (core) [alt=<6>{fill=GenericHighlight}{}] {Uberon};
-        \node (fbbt) [alt=<{1,7}>{fill=GenericHighlight}{}] at (150:2cm) {FBbt};
-        \node (zfa)  [alt=<{2,7}>{fill=GenericHighlight}{}] at (90:2cm)  {ZFA};
-        \node (xao)  [alt=<{3,7}>{fill=GenericHighlight}{}] at (30:2cm)  {XAO};
-        \node (ma)   [alt=<{4,7}>{fill=GenericHighlight}{}] at (210:2cm) {MA};
-        \node (wbbt) [alt=<{5,7}>{fill=GenericHighlight}{}] at (270:2cm) {WBbt};
+        \node (core) [alt=<6| handout:0>{fill=GenericHighlight}{}] {Uberon};
+        \node (fbbt) [alt=<{1,7| handout:0}>{fill=GenericHighlight}{}] at (150:2cm) {FBbt};
+        \node (zfa)  [alt=<{2,7| handout:0}>{fill=GenericHighlight}{}] at (90:2cm)  {ZFA};
+        \node (xao)  [alt=<{3,7| handout:0}>{fill=GenericHighlight}{}] at (30:2cm)  {XAO};
+        \node (ma)   [alt=<{4,7| handout:0}>{fill=GenericHighlight}{}] at (210:2cm) {MA};
+        \node (wbbt) [alt=<{5,7| handout:0}>{fill=GenericHighlight}{}] at (270:2cm) {WBbt};
         \node (tx1) at  (310:2cm) {...};
@@
         \only<1>{\emph{Drosophila} Anatomy Ontology}
-        \only<2>{Zebrafish Anatomy Ontology}
-        \only<3>{Xenopus Anatomy Ontology}
-        \only<4>{Mouse Anatomy Ontology}
-        \only<5>{\emph{C. elegans} Anatomy Ontology}
+        \only<2| handout:0>{Zebrafish Anatomy Ontology}
+        \only<3| handout:0>{Xenopus Anatomy Ontology}
+        \only<4| handout:0>{Mouse Anatomy Ontology}
+        \only<5| handout:0>{\emph{C. elegans} Anatomy Ontology}
       \end{overlayarea}
@@
 
-  \only<1>{
+  \only<1| handout:1>{
     \begin{columns}[t]
@@
   }
-  \only<2>{
+  \only<2| handout:2>{
     \begin{columns}
@@
   \begin{overlayarea}{\textwidth}{3cm}
-    \only<1-2>{
+    \only<1-2| handout:1>{
       \begin{block}{Inverting a mapping}\ttfamily\scriptsize
@@
     }
-    \only<3-4>{
+    \only<3-4| handout:2>{
       \begin{block}{Generating a simple axiom}\ttfamily\scriptsize
@@
     }
-    \only<5-10>{
+    \only<5-10| handout:3>{
       \begin{block}{Generating a more complex axiom}\ttfamily\scriptsize
@@
   \begin{overlayarea}{\textwidth}{2.5cm}
-    \only<1-2>{\textcolor<1>{red}{Select mappings with a UBERON subject}
+    \only<1-2| handout:1>{\textcolor<1>{red}{Select mappings with a UBERON subject}
 
     and \textcolor<2>{red}{flip them around the predicate}}
-    \only<3-4>{\textcolor<3>{red}{Select mappings with a \emph{skos:broadMatch} predicate}
+    \only<3-4| handout:2>{\textcolor<3>{red}{Select mappings with a \emph{skos:broadMatch} predicate}
 
     and \textcolor<4>{red}{create a \emph{SubClassOf} axiom}}
-    \only<5-10>{Select mappings
+    \only<5-10| handout:3>{Select mappings
       \begin{itemize}
@@
 \appendix
-%\section{\appendixname}
+\begin{frame}[noframenumbering]
+  \frametitle{Questions?}
+\end{frame}
 

### 2024-07-01 presentation.tex
Small fixes to presentation text.

@@
 
-  \note{In response to the limits of the previous system, here are the key
-  principles that dictated the design of the new approach.
+  \note{In response to the limits of the previous system, we set out to design
+  a new system, with the following aims:
 
@@
 
-  \note<1>{Over the past three years, independently of Uberon, a new standard has
-  been devised, specifically for the purpose of representing semantic
-  mappings: the Simple Standard for Sharing Ontological Mappings.
+  \note<1>{Over the past three years, independently of Uberon, a new standard
+  has been devised to work with semantic mappings: the Simple Standard for
+  Sharing Ontological Mappings.
 
-  The SSSOM standard defines a common data model (formalised in LinkML) to
-  represent and manipulate mappings in programming languages. The model
-  notably allows to capture a rich set of metadata about each mapping.
+  The SSSOM standard defines a common data model to represent and manipulate
+  mappings in programming languages. The model notably allows to capture a
+  rich set of metadata about each mapping.
 
@@
   exchange mappings, the most important of which being the ``SSSOM/TSV''
-  format, which as the name suggests is based on the ubiquitous TSV format,
-  and is intended to make reusing mappings across resources as easy as
+  format, intended to make reusing mappings across resources as easy as
   possible.
@@
   looks like: one mapping per line, each mapping consisting, at its core, of a
-  subject, predicate, object triple, associated with several metadata.}
+  triple made of a subject, an object, and a mapping predicate. That core is
+  associated with several metadata.}
 
@@
 
-  \note{One of the benefits of the SSSOM standard is that it does not simply
-  allow to say that two terms are mapped, to each other.  As we've seen in the
-  last example, it also allows to specify an explicit mapping predicate, so as
-  to give a precise meaning to each mapping. It is then possible to
-  distinguish, for example between mappings of terms that represent exactly
-  the same concept in two different vocabularies, and mappings of terms that
-  represent \emph{similar}, but not exactly identical, concepts.
+  \note{One of the benefits of the SSSOM standard is that it allows to specify
+  an explicit mapping predicate, so as to give a precise meaning to each
+  mapping. It is then possible to distinguish, for example between mappings of
+  terms that represent exactly the same concept in two different vocabularies,
+  and mappings of terms that represent \emph{similar}, but not exactly
+  identical, concepts.
 
@@
   \emph{Semantic Mapping Vocabulary}, to host, among other things, various
-  mapping predicates that can be used to describe mappings. In that
-  vocabulary, as part of this project, we have added a handful of predicates
-  to specifically represent \emph{cross-species mappings}.
+  mapping predicates that can be used to qualify mappings. In that vocabulary,
+  as part of this project, we have added a handful of predicates to
+  specifically represent \emph{cross-species mappings}.
 
@@
 
-  \note{Now that we can store mappings in a dedicated format, we need a way
-  to use those mappings when we want to integrate several ontologies together.
-  And, as per our third goal, we want to do that in a way that is, as much as
-  possible, not specific to Uberon, but reusable by other projects as well.
+  \note{Now that we can store mappings in a dedicated format, along with their
+  metadata and with meaningful mapping predicates, we need a way to use those
+  mappings when we want to integrate several ontologies together.  And, as per
+  our third goal, we want to do that in a way that is, as much as possible,
+  not specific to Uberon, but reusable by other projects as well.
 
@@
 
-  The mappings are transformed at build-time into OWL bridge files using a
-  custom SSSOM/T-OWL ruleset, that contains rules similar to the one we have
-  seen earlier: a rule that says that cross-species mappings between FBbt and
-  Uberon are to be transformed into equivalence axioms with a taxonomic
-  restriction to the \emph{Drosophila melanogaster} taxon, a rule that says
-  that cross-species mappings between XAO and Uberon are to be transformed
-  into equivalence axioms with a taxonomic restriction to the \emph{Xenopus}
-  taxon, etc.
-
-  That step is performed by a dedicated ROBOT plugins that implement the
-  SSSOM/T-OWL language.
+  The bridge files are generated by a ROBOT plugin that implements the
+  SSSOM/T-OWL language, and a custom SSSOM/T-OWL ruleset that contains the
+  Uberon-specific logic, in the form of rules similar to the one seen in the
+  last example.
 
@@
   standard ROBOT merge operation, producing the multi-species ontology that we
-  want.}
+  want.
+
+  Overall, that pipeline is both more robust, as most of the code is
+  consolidated in ROBOT, and more flexible, as the Uberon-specific logic is
+  easily modifiable in the SSSOM/T-OWL ruleset.}
 
@@
 
-  \note{For Uberon itself, the main benefit of the new approach is that we now
-  have a pipeline that is both more flexible --~for example, easy to modify,
-  as all the logic is in the SSSOM/T-OWL ruleset~-- and more robust.
+  \note{Beyond the more robust and more flexible pipeline, the new approach
+  benefits for downstream users of Uberon as well.
 
-  For downstream users of Uberon, the mappings between Uberon and the
-  taxon-specific ontologies are now first-class artefacts. They can have their
-  own metadata, allowing for traceability and accountability of each
-  individual mapping statement. They can be maintained separately from the
-  Uberon ontology itself, for example to delegate them to the taxon-specific
-  ontologies themselves (something we have already started doing for the
-  mappings with FBbt, which are now maintained by FBbt rather than by Uberon).
-  They are available as a distinct release product, which is findable on its
-  own, in a format that is still relatively new but that we hope will become
-  the standard format for semantic mappings.
+  The mappings between Uberon and the taxon-specific ontologies are now
+  first-class artefacts. They can have their own metadata, allowing for
+  traceability and accountability of each individual mapping statement. They
+  can be maintained separately from the Uberon ontology itself, for example to
+  delegate them to the taxon-specific ontologies themselves (something we have
+  already started doing for the mappings with FBbt, which are now maintained
+  by FBbt rather than by Uberon).  They are available as a distinct release
+  product, which is findable on its own, in a format that is still relatively
+  new but that we hope will become the standard format for semantic mappings.
 

### 2024-07-01 presentation.tex
Add back removed slides about the previous system.
We are now aiming for a 25-min talk, so we can afford to put back the
previously removed slides that expand about the limitatons of the
previous integration pipeline (but put the slide with the overview
first, not last).

Also move the slide about the ROBOT implementation back from the
supplementary slides to the main presentation.

@@
     \tableofcontents[currentsection]
+  \end{frame}
+}
+\AtBeginSubsection{
+  \begin{frame}{Outline}{}
+    \tableofcontents[currentsection,currentsubsection]
   \end{frame}
@@
 
-\section{Limits of the previous system}
+\section{Existing system and its limitations}
 
 \begin{frame}
-  \frametitle{Overview of the previous integration system}
+  \frametitle{Integration pipeline overview}
 
@@
         \node [Entity,text width=1.3cm,alt=<1>{fill=GenericHighlight}{}] (uberon) {Uberon\\(incl. xrefs)};
-        \node [Process,below right=of uberon,alt=<3>{fill=GenericHighlight}{}] (script) {Custom script};
-        \node [Entity,above right=of script,align=left] (ssao) {Taxon-specific\\
-                                                                ontologies};
+        \node [Process,below right=of uberon] (script) {Custom script};
+        \node [Entity,above right=of script,align=left] (ssao) {Taxon-specific\\ontologies};
         \node [Entity,below=of script,alt=<2>{fill=GenericHighlight}{}] (cob) {Bridge files};
@@
     \end{column}
-    \begin{column}{.51\textwidth}
+    \begin{column}{.53\textwidth}
       \begin{block}{Mappings maintained as \textcolor<1>{red}{xrefs}}\scriptsize\ttfamily
@@
 
+  \note<1>{Here is a brief overview of the integration pipeline.
+
+  Mappings between Uberon and the taxon-specific ontologies are maintained as
+  cross-references that are carried by the taxon-neutral terms in Uberon and
+  that point to the taxon-specific terms in other ontologies -- as shown here,
+  where Uberon's ``dorsal vessel heart'' term carries a ``xref'' to
+  FBbt:0003154, which is FBbt's term for the fly heart.
+
+  At build time, xrefs are extracted by a custom script that turns them into
+  bridge files, which contains OBO materialisations of the mappings, in the
+  form of equivalence axioms (such as an axiom stating that FBbt's ``heart''
+  is equivalent to a Uberon ``dorsal vessel heart'' that is part of the
+  \emph{Drosophila melanogaster} taxon).
+
+  While the process looks superficially simple (though it is, in fact, quite
+  complex, it's just that most of the complexity is hidden behind that
+  script), it is also severely limited.
+
+  The limitations stem both from the way the mappings are represented, and the
+  way they are transformed into bridge files.}
+
+\end{frame}
+
+\begin{frame}
+  \frametitle{Issues with mappings maintained as cross-references}
+
+  \begin{block}{Mappings embedded within the ontology}
     \begin{itemize}
-    \item Mappings not available to third parties
-    \item \textcolor<3>{red}{Custom pipeline} highly Uberon-specific, inflexible and not reusable
+      \item Not available separately
      \end{itemize}
+  \end{block}
 
-  \note<1>{Mappings are represented and maintained as \emph{cross-references}
-  that are carried by the taxon-neutral terms in Uberon and that point to the
-  taxon-specific terms in other ontologies -- as shown here, where Uberon's
-  ``dorsal vessel heart'' term carries a ``xref'' to FBbt:00003154, which is
-  FBbt's term for the fly heart.
+  \begin{block}{No additional data}
+    \begin{itemize}
+      \item All the cross-references say is that the mappings exist
+    \end{itemize}
+  \end{block}
 
-  When building Uberon, those cross-references are extracted by a custom
-  script that turns them into \emph{bridge files}, which contains OWL
-  materialisations of the mappings, in the form of equivalence axioms (for
-  example here, an axiom stating the FBbt's ``heart'' is equivalent to an
-  Uberon's ``dorsal vessel heart'' that is part of the \emph{Drosophila
-  melanogaster} taxon).
+  \begin{block}{Meaningless mapping predicate}
+    \begin{itemize}
+      \item Overloading of the \emph{oboInOwl:hasDbXref} property
+      \item Workaround: {\ttfamily\scriptsize treat-xrefs-as-reverse-genus-differentia FBbt NCBITaxon:7227}
+        \begin{itemize}
+          \item Assigns a meaning to \emph{all} cross-references between FBbt and Uberon
+        \end{itemize}
+    \end{itemize}
+  \end{block}
 
-  This system has severe limitations, one of the most important being that the
-  mappings are directly embedded within the ontology itself and are not
-  independently available. In FAIR terms, this makes the mappings much more
-  difficult to find, access and reuse than we would like.
+  \note{The one benefit of using xrefs to represent mappings is that it is a
+  very simple method, but that simplicity comes at a high cost.
 
-  Furthermore, the custom script that generates the bridge files from the
-  cross-references is highly specific to Uberon; its behaviour are completely
-  hardcoded to fit Uberon's use case, so the integration logic is difficult to
-  modify and almost impossible to reuse by other projects.}
+  First, the mappings are embedded within the ontology and as such are not
+  independently available -- they need to be forcefully extracted whenever
+  someone needs them. This makes the mappings more difficult to find, access,
+  and reuse than it should be.
+
+  Second, the cross-reference is telling us that there is a mapping, but we
+  can't know anything else about that mapping. Why does that mapping exist?
+  Who asserted it? When? On what basis? This is simply not recorded.
+
+  Third, the \emph{oboInOwl:hasDbXref} annotation has been used for a lot of
+  things since its introduction, and no longer has a clear and unambiguous
+  meaning -- what it means depends on where it is used. At best, it indicates
+  that two terms are mapped \emph{somehow}, but the nature of the relation is
+  unspecified.
+
+  In Uberon we can circumvent that last issue by using supplementary
+  ontology-level annotations to provide meaning to some cross-references. But
+  this is more a workaround than a proper solution, and it comes with problems
+  of its own.}
+
+\end{frame}
+
+\begin{frame}
+  \frametitle{Issues with the bridge generation process}
+
+  \begin{columns}
+    \begin{column}{.4\textwidth}
+      \begin{center}
+        \begin{tikzpicture}
+          \node [File] (uberon) at (0,0)    {uberon.obo};
+          \node [File] (script) at (0,-1)   {make-bridge-ontology-from-xrefs.pl};
+          \node [File] (bridge) at (0,-2)   {uberon-bridge-to-fbbt.obo};
+          \node [File]          at (0,-2.5) {uberon-bridge-to-wbbt.obo};
+          \node [File]          at (0,-3)   {uberon-bridge-to-ma.obo};
+          \node [File]          at (0,-3.5) {uberon-bridge-to-....obo};
+          \draw[->] (uberon) -- (script) (script) -- (bridge);
+        \end{tikzpicture}
+      \end{center}
+    \end{column}
+    \begin{column}{.55\textwidth}
+      \begin{itemize}
+        \item Lack of flexibility
+        \item Maintenance nightmare
+        \item Dependent on the OBO format
+        \item Highly Uberon-specific
+      \end{itemize}
+    \end{column}
+  \end{columns}
+
+  \note{The bridge files are generated by this custom Perl script, which reads
+  the Uberon ontology file and generates one bridge file for each
+  taxon-specific ontology. That script is plagued with several issues.
+
+  The process is very rigid. For example, it is impossible to generate
+  only one bridge file on-demand. Almost everything is hardcoded in the
+  script, making it difficult to change anything whenever needed.
+
+  The script is difficult to maintain for most Uberon editors and developers,
+  in no small part because it conflates the technical aspects of extracting
+  the cross-references and generating the bridges, and the logical aspects of
+  determining which bridging axioms should be produced.
+
+  It's a smaller issue, but the script also only works with the OBO files,
+  both for its input and its output.
+
+  But the main problem is that the bridge generation script is highly specific
+  to Uberon. It is almost impossible to reuse its logic elsewhere. Yet the
+  need for integrating several ontologies together, using mappings as glue to
+  connect terms between those ontologies, is certainly not a Uberon-specific
+  need.}
 
@@
   \begin{enumerate}
-    \item Cross-ontology mappings (COM) as first-class entities
+    \item Treat cross-ontology mappings as proper entities
       \begin{itemize}
-        \item available independently of the ontology itself
+        \item Mappings set made available as independent artefact
       \end{itemize}
-    \item Represent mappings using a standard format intended for that purpose
+    \item Represent COMs using a standard format intended for that purpose
       \begin{itemize}
@@
       \end{itemize}
-    \item Make it easy to derive OWL bridge files from mappings
+    \item Make it easy to derive bridging axioms from mappings
       \begin{itemize}
@@
 
-  First, we want to treat mappings as ``first-class entities'', on the same
-  level as the terms of the ontology. This notably means that they must be
-  publicly available on their own, without requiring people interested in them
-  to forcefully extract them from the ontology.
+  First, we want the cross-ontology mappings to be treated as first-class
+  objects, similarly to the terms of the ontology, rather than merely
+  annotations on the terms. They should in fact be treated as the source
+  artefact from which the bridge files are derived. As with other source
+  artefacts, they must be published on their own and be independently findable.
 
@@
   it possible to store mapping-specific metadata, and to give precise meaning
-  to a mapping (not just saying that two terms are mapped, but: who asserted
-  that those two terms should be mapped? When? On what basis? And what does
-  the mapping actually means?).
+  to every individual mapping (everything that is not possible with the
+  cross-reference format).
 
@@
 
-\section{Implementation}
+\section{Implementation of a new approach}
+
+\subsection{New representation of mappings}
 
@@
 \begin{frame}
-  \frametitle{New mapping predicates for cross-species mappings}
+  \frametitle{Meaningful predicates for cross-species mappings}
 
@@
 
-  \note{One of the benefits of the SSSOM standard is that it allows to specify
+  \note{One of the benefits of the SSSOM model is that it allows to specify
   an explicit mapping predicate, so as to give a precise meaning to each
@@
 \end{frame}
+
+\subsection{New bridge generation process}
 
@@
   First, a simple rule with a simple filter that selects all mappings where
-  the subject is a Uberon term, and apply an inversion preprocessor to them.
+  the subject is a Uberon term, and applies an inversion preprocessor to them.
   Inversion flips a mapping around the predicate, so that the subject becomes
@@
 \begin{frame}
+  \frametitle{Adding SSSOM/T-OWL support to ROBOT}
+
+  \begin{block}{Prerequisite step}
+    Support for ``pluggable commands'' in ROBOT (since version 1.9.5)
+  \end{block}
+
+  \begin{block}{sssom:inject command}
+    Apply \textcolor<2>{red}{SSSOM/T-OWL} rules to \textcolor<3>{red}{a
+    mapping set} and inject the resulting axioms into the
+    \textcolor<4>{red}{current ontology}
+
+    \medskip
+    {\ttfamily\scriptsize
+robot sssom:inject -i \textcolor<4>{red}{input.owl} --sssom \textcolor<3>{red}{mappings.sssom.tsv} --ruleset \textcolor<2>{red}{bridge.rules}
+    }
+  \end{block}
+
+  \note<1>{With the SSSOM/T-OWL language fully designed, we then needed to
+  actually implement it. We chose to add support for SSSOM/T-OWL to ROBOT,
+  since this is the main tool we use for most of our pipelines.
+
+  The first step was to add plugin support for ROBOT -- that is, the
+  possibility to add third-party commands to ROBOT. This is so that we could
+  quickly add new commands, such as SSSOM-related commands, to ROBOT without
+  being tied by ROBOT's release cycle. This was done with version 1.9.5 of
+  ROBOT, released at the end of Summer 2023.
+
+  We then developed a SSSOM plugin for ROBOT, which among other things,
+  provides a command that implement our SSSOM/T-OWL language. That command
+  will read a set of SSSOM/T-OWL rules and apply them to a set of mappings,
+  thereby producing bridging axioms which will then be injected into the
+  ontology that is currently being manipulated by ROBOT.}
+
+\end{frame}
+
+\begin{frame}
   \frametitle{New SSSOM-based integration pipeline}
@@
           \node [Entity] (uberon) {Uberon};
-          \node [Entity,right=of uberon] (com) {Mapping sets};
-          \node [Entity,right=of com,align=left] (ssao) {Taxon-specific\\
+          \node [Entity,right=of uberon,alt=<2>{fill=GenericHighlight}{}] (com) {Mapping sets};
+          \node [Entity,right=of com,align=left,alt=<3>{fill=GenericHighlight}{}] (rules) {SSSOM/T-OWL\\
+                                                                                           ruleset};
+          \node [Process,below=of com,alt=<4>{fill=GenericHighlight}{}] (inject) {sssom:inject};
+          \node [Process,below=of inject,alt=<5>{fill=GenericHighlight}{}] (merge) {merge};
+          \node [Entity,right=of merge,align=left] (ssao) {Taxon-specific\\
                                                          ontologies};
-          \node [Process,below=of com,alt=<2>{fill=GenericHighlight}{}] (sssomt) {SSSOM/T-OWL};
-          \node [Entity,below=of sssomt] (cob) {Bridge files};
-          \node [Process,below=of cob,alt=<3>{fill=GenericHighlight}{}] (merge) {OWL merge};
           \node [Entity,below=of merge] (cm) {Multi-species ontology};
-          \draw [->] (com) -- (sssomt) (sssomt) -- (cob);
-          \draw [->] (cob) -- (merge) (merge) -- (cm);
-          \draw [->] (uberon.south) |- (merge.west);
-          \draw [->] (ssao.south) |- (merge.east);
+          \draw [->] (com) -- (inject) (inject) -- (merge) (merge) -- (cm);
+          \draw [->] (rules.south) |- (inject.east);
+          \draw [->] (uberon.south) |- (inject.west);
+          \draw [->] (ssao.west) -- (merge.east);
         \end{tikzpicture}
@@
       \begin{enumerate}
-        \item Generating the bridge files through a \textcolor<2>{red}{SSSOM/T-OWL ruleset}\\
-              (\texttt{robot sssom:inject})
-        \item \textcolor<3>{red}{Merging} Uberon, the bridges, and the taxon-specific ontologies\\
-              (\texttt{robot merge})
+        \item Maintaining (and collecting) mappings as \textcolor<2>{red}{SSSOM sets}
+        \item Maintaining the integration logic as \textcolor<3>{red}{SSSOM/T-OWL rules}
+        \item Generating the bridging axioms through \textcolor<4>{red}{SSSOM/T-OWL support in ROBOT}
+        \item \textcolor<5>{red}{Merging in} the taxon-specific ontologies
       \end{enumerate}
@@
 
-  \note<1>{This leads us to our new, SSSOM-based integration pipeline for Uberon.
-  That pipeline has three distinct inputs: (1) Uberon itself, (2) the mappings
-  between Uberon and the taxon-specific ontologies (which are no longer
-  embedded within the ontology itself), and (3) the taxon-specific ontologies.
+  \note<1>{This leads us to our new, SSSOM-based integration pipeline for
+  Uberon.
 
-  The bridge files are generated by a ROBOT plugin that implements the
-  SSSOM/T-OWL language, and a custom SSSOM/T-OWL ruleset that contains the
-  Uberon-specific logic, in the form of rules similar to the one seen in the
-  last example.
+  The mappings are now maintained separately from Uberon itself, as SSSOM
+  mapping sets.
 
-  After this step, we have only OWL files, which can all be merged using a
-  standard ROBOT merge operation, producing the multi-species ontology that we
-  want.
+  The integration logic, which decides how the bridging axioms should be
+  generated, is also maintained separately, as a large ruleset written in the
+  SSSOM/T-OWL language. It contains rules similar to the ``complex'' example
+  of rule shown previously.
+
+  At build time, we use our SSSOM ROBOT plugin to apply that ruleset to our
+  mappings, and inject the resulting bridging axioms into Uberon.
+
+  Then, we just need to merge the taxon-specific ontologies into the result of
+  the previous step, producing the multi-species ontology that we want.
 
   Overall, that pipeline is both more robust, as most of the code is
-  consolidated in ROBOT, and more flexible, as the Uberon-specific logic is
-  easily modifiable in the SSSOM/T-OWL ruleset.}
+  consolidated in the ROBOT plugin, and more flexible, as the decision logic
+  is easily modifiable in the SSSOM/T-OWL ruleset.}
 
@@
 
-  \note{Beyond the more robust and more flexible pipeline, the new approach
-  benefits for downstream users of Uberon as well.
+  \note{Beyond the more robust and more flexible pipeline, which is mostly of
+  interest for Uberon editors and developers, the new approch has broader
+  benefits.
 
@@
   first-class artefacts. They can have their own metadata, allowing for
-  traceability and accountability of each individual mapping statement. They
-  can be maintained separately from the Uberon ontology itself, for example to
-  delegate them to the taxon-specific ontologies themselves (something we have
-  already started doing for the mappings with FBbt, which are now maintained
-  by FBbt rather than by Uberon).  They are available as a distinct release
+  traceability and accountability of each individual mapping statement, and
+  their own mapping predicate. They can be maintained separately from the
+  Uberon ontology itself, and we can even delegate the maintenance of some of
+  them to the taxon-specific ontologies themselves (something we have already
+  started doing for the mappings with FBbt, which are now maintained by
+  FlyBase rather than by Uberon).  They are available as a distinct release
   product, which is findable on its own, in a format that is still relatively
@@
 
-  And for everyone else, we now have a generic framework to integrate multiple
+  And beyond Uberon, we now have a generic framework to integrate multiple
   ontologies with SSSOM-maintained mappings. The SSSOM/T-OWL language, and its
@@
   tailored to the needs of Uberon.}
+
+\end{frame}
+
+\begin{frame}[fragile]
+  \frametitle{Example of reuse: FBbt/Uberon integration test}
+
+  A single ROBOT command to \textcolor<1>{red}{merge FBbt with Uberon},
+  \textcolor<2>{red}{inject the mappings-derived bridging axioms}, and
+  \textcolor<3>{red}{check for possible inconsistencies} that would reveal
+  incorrect mappings:
+
+  \begin{block}{}
+\begin{semiverbatim}
+robot \textcolor<1>{red}{merge} -i \textcolor<1>{red}{fbbt.owl} -i \textcolor<1>{red}{uberon.owl} \\
+      \textcolor<2>{red}{sssom:inject} --sssom \textcolor<2>{red}{mappings.sssom.tsv}   \\
+                   --ruleset \textcolor<2>{red}{bridging.rules}     \\
+      \textcolor<3>{red}{reason} -r ELK
+\end{semiverbatim}
+  \end{block}
+
+  \note<1>{Lastly, here is an example to illustrate some of the benefits of
+  reusable integration logic.
+
+  Because FBbt and Uberon are developed separately, there is always a risk
+  that, whenever something changes in either FBbt or Uberon, an inter-ontology
+  inconsistency is introduced.  Because the logic to integrate FBbt into
+  Uberon was impossible to reuse outside of Uberon, it was difficult for FBbt
+  editors to detect and prevent such inconsistencies whenever we edited the
+  ontology.
+
+  But with the SSSOM approach, such testing can be done with a single ROBOT
+  command in three steps, as shown here: (1) we merge the current versions of
+  FBbt and Uberon; (2) we derive OWL axioms from the FBbt-to-Uberon mappings
+  and inject them into the merged ontology; (3) we reason over the merged
+  ontology to detect inconsistencies. This takes but a couple of minutes, and
+  that check is now part of our test suite in FBbt. Thanks to this approach,
+  over the past two years we have rooted out all existing inconsistencies
+  between FBbt and Uberon.}
 
@@
 
-\begin{frame}[noframenumbering]
-  \frametitle{Implementation in ROBOT commands}
+\end{document}
 
-  \begin{block}{sssom:xref-extract command}
-    Convert existing mappings represented as cross-references into a SSSOM set
-
-    \medskip
-    {\ttfamily\scriptsize
-robot sssom:xref-extract -i input.owl --mapping-file output.sssom.tsv
-    }
-  \end{block}
-
-  \begin{block}{sssom:inject command}
-    Apply SSSOM/T-OWL rules to a mapping set and inject the resulting axioms into an ontology
-
-    \medskip
-    {\ttfamily\scriptsize
-robot sssom:inject -i input.owl --sssom mappings.sssom.tsv --ruleset bridge.rules
-    }
-  \end{block}
-
-  \note{To actually implement our new approach, we first, as a preliminary
-  step, added plugin support to ROBOT, the main tool we use to manipulate OWL
-  and OBO ontologies. This is so that we can quickly add new commands to
-  ROBOT. We then developed a SSSOM plugin for ROBOT, which provides two
-  distinct commands.
-
-  The first is a command to facilitate the transition to SSSOM-maintained
-  mappings by automatically extracting mappings stored as cross-references
-  within an ontology into a mapping set in the SSSOM/TSV format.
-
-  More importantly, the second is a command that implements the SSSOM/T-OWL
-  domain-specific language. That command takes a mapping set and a set of
-  SSSOM/T-OWL rules as input, and will inject into the current ontology the
-  axioms produced by applying the SSSOM/T-OWL rules to the mapping set.}
-
-\end{frame}
-
-\begin{frame}[fragile,noframenumbering]
-  \frametitle{Example: FBbt/Uberon/CL integration test}
-
-  A single ROBOT command to merge FBbt with Uberon, CL, and the mappings-derived
-  bridging axioms, and check for possible inconsistencies that would reveal
-  incorrect mappings:
-
-  \begin{block}{}
-\begin{verbatim}
-robot merge -i fbbt.owl -i uberon.owl -i cl.owl \
-      sssom:inject --sssom mappings.sssom.tsv   \
-                   --ruleset bridging.rules     \
-      reason -r ELK
-\end{verbatim}
-  \end{block}
-
-  \note{Lastly, here is an example to illustrate some of the benefits of the
-  approach for third-party users. As I've mentioned, at FlyBase we are now
-  maintaining ourselves the mappings between FBbt and Uberon.
-
-  Because FBbt and Uberon are developed separately, there is always a risk
-  that, whenever something changes in either FBbt, Uberon, or the
-  FBbt-to-Uberon mappings, an inter-ontology inconsistency is introduced.
-  Testing to detect and prevent such inconsistencies is difficult and,
-  historically, has never actually been done on a regular basis.
-
-  But with the SSSOM approach, such testing can be done with a single ROBOT
-  command in three steps, as shown here: (1) we merge the current versions of
-  FBbt, Uberon, and CL; (2) we derive OWL axioms from the FBbt-to-Uberon/CL
-  mappings and inject them into the merged ontology; (3) we reason over the
-  merged ontology to detect inconsistencies. This takes but a couple of
-  minutes, and that test is now part of our test suite in FBbt. Thanks to this
-  approach, over the past two years we have rooted out all existing
-  inconsistencies between FBbt and Uberon.}
-
-\end{frame}
-
-\end{document}
+% vim: tw=78 et ai ts=2 st=2 sw=2
